### 2024-11-26 paper/paper.qmd
Initial Commiit + Update Script

---
title: "2025 US School Shooting Predictions and Solutions"
subtitle: "Leveraging Forecasting to Shape Policy and Protect Communities"
author: 
  - Yun Chu
thanks: "Code and data are available at: https://github.com/chuyun2024/School-Shooting-Analysis."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)

#### Read data ####
data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

Paragraph of Estimand!

# Data {#sec-data}

## Overview

The dataset has 416 entries, with each entry representing a unique school shooting incident. Incidents occurring during after-hours events, accidental gun discharges that only injured the individual handling the firearm, and private suicides that did not endanger other children were excluded from consideration. Additionally, shootings at colleges and universities, which involve young adults rather than children, were not included in the analysis [@washingtonpost_school_shootings]. These entries cover 50 variables that provide information about the schools, locations, date and time of shooting, shooters details, number of killed and injured, and the relationship of the shooter to school, the weapon type and source.

As the federal government does not consistently track school shootings, this dataset from *The Washington Post* fills a critical gap. It was carefully assembled using information from diverse sources, including news articles, open-source databases, law enforcement reports, and direct inquiries to schools and police departments. Although sources like FBI crime reports and local school incident logs were reviewed, they lack the detail and comprehensive coverage of this dataset. Its unparalleled breadth and depth make it the strongest foundation for predictive modeling and generating actionable insights.

We use the statistical programming language R [@citeR] to download, clean, analyze and model the US School Shooting Data. The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] . The following libraries are utilized in this paper:

- tidyverse [@tidyverse2024]
- dplyr [@dplyr2024]
- lubridate [@lubridate2024]
- readr [@readr2024]
- stringr [@stringr2024]
- arrow [@arrow2024]
- testthat [@testthat2024]
- randomForest [@randomForest2024]
- ggplot2
- maps

## Measurement

The dataset, compiled by The Washington Post, translates real-world school shooting incidents into structured entries by aggregating information from news articles, open-source databases, law enforcement reports, and direct calls to schools. Only verified incidents, such as shootings during school hours or on school property, were included. Events like after-hours shootings, private suicides, or accidental discharges without other injuries were excluded [@washingtonpost_school_shootings].

## Summary Statistics & Relationship Between Variables

$ Causalities = Killed + Injured $

## Outcome variables

The outcome variable for this analysis is number of causalities for each state in 2025.

@fig-Number_of_Incidents_Over_Time shows the number of school shooting incidents in US since 1999. The number of school shooting incidents almost doubled since 2020 compared with previous years.

```{r}
#| label: fig-Number_of_Incidents_Over_Time
#| fig-cap: "Number of US School Shooting Incidents Over Time"
#| echo: false

ggplot(data, aes(x = date)) +
  geom_histogram(binwidth = 60, fill = '#3FA0FF', color = 'white', alpha = 0.8) +
  labs(
    title = "Number of US School Shooting Incidents Over Time",
    subtitle = "Visualizing school shooting trends in the US",
    x = "Date",
    y = "Incident Count"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 15)),
    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 10)),
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.grid.major = element_line(color = "#E5E5E5"),
    panel.grid.minor = element_blank()
  )

```

@fig-Shooting_Casaulties_by_State visualizes the number of school shooting incidents in US by state for all data since 1999. California, Texas, Florida and North Carolina all have more than 20 school shooting incidents in the past 25 years while other states has less than 20 school shootings.

```{r}
#| label: fig-Shooting_Casaulties_by_State
#| fig-cap: "US School Shooting Casaulties by State"
#| echo: false

# Aggregate data to calculate total casualties by state
state_counts <- data %>%
  group_by(state) %>%
  summarise(total_casualties = sum(casualties, na.rm = TRUE))

# Get US map data
us_map <- map_data("state")

# Ensure state names are in lowercase to match map_data
state_counts <- state_counts %>%
  mutate(state = tolower(state))

# Merge map data with shooting casualties data
map_data <- us_map %>%
  left_join(state_counts, by = c("region" = "state"))

# Replace NA values in `total_casualties` with 0 (for states with no recorded casualties)
map_data$total_casualties[is.na(map_data$total_casualties)] <- 0

ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",
    high = "#0072B2",
    name = "Total Casualties",
    breaks = c(0, 50, 100, 150, 200)  # Adjust the breaks based on your data
  ) +
  labs(
    title = "US School Shooting Casualties by State",
    subtitle = "Number of casualties visualized geographically",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 13, margin = ggplot2::margin(t = 10, b = 10)),
    plot.subtitle = ggplot2::element_text(hjust = 0.5, size = 10, margin = ggplot2::margin(t = 0, b = 20)),
    legend.position = "right",  # Alternative: Use "bottom" for a horizontal legend
    legend.key.height = unit(1, "cm"),  # Adjust legend size
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Increase margins for PDF export
  )

```

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: false
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: false
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: false
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-26 scripts/00-simulate_data.R

@@ -1,52 +1,122 @@
 #### Preamble ####
-# Purpose: Simulates a dataset of Australian electoral divisions, including the 
-  #state and party that won each division.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Purpose: Simulates a dataset of US School Shooting.
+# Author: Yun Chu
+# Date: 22 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
 # Pre-requisites: The `tidyverse` package must be installed
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Any other information needed? None
 
 
 #### Workspace setup ####
-library(tidyverse)
-set.seed(853)
-
+library(dplyr)
+library(lubridate)
 
 #### Simulate data ####
-# State names
-states <- c(
-  "New South Wales",
-  "Victoria",
-  "Queensland",
-  "South Australia",
-  "Western Australia",
-  "Tasmania",
-  "Northern Territory",
-  "Australian Capital Territory"
+# Set seed for reproducibility
+set.seed(123)
+
+# Number of observations to simulate
+n <- 100
+
+# Simulate 'uid' as a sequence of unique identifiers
+uid <- 1:n
+
+# Simulate 'school_name' as random school names
+school_names <- c("Lincoln High School", "Washington Middle School", "Roosevelt Elementary", "Jefferson Academy", "Madison Primary School")
+school_name <- sample(school_names, n, replace = TRUE)
+
+# Simulate 'nces_district_id' as random 7-digit numbers (as character strings)
+nces_district_id <- sprintf("%07d", sample(1e6:1e7 - 1, n, replace = TRUE))
+
+# Simulate 'district_name' as random district names
+district_names <- c("Springfield District", "Riverside District", "Mountain View District", "Lakeside District", "Green Valley District")
+district_name <- sample(district_names, n, replace = TRUE)
+
+# Simulate 'date' as random dates between 1999 and 2024
+start_date <- as.Date("1999-01-01")
+end_date <- as.Date("2024-11-22")
+date <- sample(seq(start_date, end_date, by = "day"), n, replace = TRUE)
+
[…]

### 2024-11-26 scripts/01-test_simulated_data.R

@@ -1,18 +1,15 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated Australian 
-  #electoral divisions dataset.
-# Author: Rohan Alexander
-# Date: 26 September 2024
-# Contact: rohan.alexander@utoronto.ca
-# License: MIT
+# Purpose: Tests the structure and validity of the simulated school shooting dataset.
+# Author: Yun Chu
+# Date: 22 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
 # Pre-requisites: 
-  # - The `tidyverse` package must be installed and loaded
   # - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `starter_folder` rproj
+# Any other information needed? None
 
 
 #### Workspace setup ####
-library(tidyverse)
 
 analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
 
@@ -26,64 +23,73 @@ if (exists("analysis_data")) {
 
 #### Test data ####
 
-# Check if the dataset has 151 rows
-if (nrow(analysis_data) == 151) {
-  message("Test Passed: The dataset has 151 rows.")
-} else {
-  stop("Test Failed: The dataset does not have 151 rows.")
+# Load necessary library
+library(dplyr)
+
+# Function to test the simulated dataset
+test_simulated_data <- function(data) {
+  # 1. Check the structure of the data
+  cat("\n--- Structure of the Dataset ---\n")
+  str(data)
+  
+  # 2. Check for missing values
+  cat("\n--- Missing Values ---\n")
+  missing_summary <- sapply(data, function(x) sum(is.na(x)))
+  print(missing_summary)
+  
+  # 3. Check unique values for categorical variables
+  cat("\n--- Unique Values in Categorical Variables ---\n")
+  categorical_vars <- c("school_name", "district_name", "school_type", 
+                        "shooting_type", "gender_shooter1", "shooter_relationship1")
+  for (var in categorical_vars) {
+    if (var %in% names(data)) {
+      cat("\n", var, ":\n")
+      print(table(data[[var]]))
+    }
+  }
+  
+  # 4. Check ranges for numerical variables
+  cat("\n--- Summary Statistics for Numerical Variables ---\n")
+  numerical_vars <- c("enrollment", "killed", "injured", "casualties", "age_shooter1")
+  for (var in numerical_vars) {
[…]

### 2024-11-26 scripts/02-download_data.R

@@ -1,26 +1,22 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Downloads and saves the data from the Washington Post's Github
+# Author: Yun Chu
+# Date: 22 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
+# Pre-requisites: None
+# Any other information needed? None
 
 
 #### Workspace setup ####
-library(opendatatoronto)
+
 library(tidyverse)
-# [...UPDATE THIS...]
+library(readr)
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
+school_shoot_raw_data <- read.csv("https://raw.githubusercontent.com/washingtonpost/data-school-shootings/master/school-shootings-data.csv")
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+write_csv(school_shoot_raw_data, "data/01-raw_data/raw_data.csv") 
 
          

### 2024-11-26 scripts/03-clean_data.R

@@ -1,44 +1,134 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Cleans the US School Shooting data by coverting datatypes to the right ones and extracting each
+# component of date columns
+# Author: Yun Chu
+# Date: 22 November 2022
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
+# Pre-requisites: 
+# - 02-downloaded_data.R must have been run
+# Any other information needed? None
 
 #### Workspace setup ####
-library(tidyverse)
+library(dplyr)
+library(lubridate)
+library(stringr)
+library(arrow)
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
+# Load the dataset
+school_shoot_raw_data <- read.csv("data/01-raw_data/raw_data.csv")
+
+# Column explanations from the provided file
+column_explanations <- list(
+  uid = "Unique identifier",
+  nces_school_id = "National Center for Education Statistics unique identifier for the school",
+  school_name = "Name of the school",
+  nces_district_id = "National Center for Education Statistics unique identifier for the school district",
+  district_name = "Name of the school district",
+  city = "City where the school is located",
+  state = "State where the school is located",
+  enrollment = "Total enrollment of the school",
+  killed = "Number of individuals killed during the incident",
+  injured = "Number of individuals injured during the incident",
+  date = "Date of the incident"
+)
+
+# Step 1: Inspect the dataset
+str(school_shoot_raw_data)
+summary(school_shoot_raw_data)
+head(school_shoot_raw_data)
+
+# Step 2: Rename columns for consistency
+school_shoot_cleaned <- school_shoot_raw_data %>%
+  rename(
+    uid = uid,
+    nces_school_id = nces_school_id,
[…]

### 2024-11-26 scripts/04-test_analysis_data.R

@@ -1,69 +1,115 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 26 September 2024 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Tests cleaned school shoots data
+# Author: Yun Chu
+# Date: 22 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
+# Pre-requisites: 
+# 03-clean_data.R has been run
+# Any other information needed? None
 
 
 #### Workspace setup ####
-library(tidyverse)
+library(dplyr)
 library(testthat)
+library(arrow)
 
 data <- read_csv("data/02-analysis_data/analysis_data.csv")
 
 
 #### Test data ####
-# Test that the dataset has 151 rows - there are 151 divisions in Australia
-test_that("dataset has 151 rows", {
-  expect_equal(nrow(analysis_data), 151)
-})
 
-# Test that the dataset has 3 columns
-test_that("dataset has 3 columns", {
-  expect_equal(ncol(analysis_data), 3)
-})
+# Function to test cleaned dataset
+test_cleaned_data <- function(data) {
+  cat("\n--- Testing Cleaned Dataset ---\n")
+  
+  # 1. Check the structure of the dataset
+  cat("\n--- Dataset Structure ---\n")
+  print(str(data))
+  
+  # 2. Check for missing values
+  cat("\n--- Missing Values ---\n")
+  missing_values <- sapply(data, function(x) sum(is.na(x)))
+  print(missing_values)
+  
+  # 3. Validate data types for key columns
+  cat("\n--- Data Type Validation ---\n")
+  expected_types <- list(
+    uid = "character",
+    nces_school_id = "character",
+    school_name = "character",
+    nces_district_id = "character",
+    district_name = "character",
+    date = "Date",
+    enrollment = "numeric",
+    killed = "numeric",
+    injured = "numeric",
+    casualties = "numeric"
+  )
+  for (col in names(expected_types)) {
+    if (col %in% names(data)) {
+      actual_type <- class(data[[col]])[1]
[…]

### 2024-11-26 scripts/05-exploratory_data_analysis.R

@@ -1,37 +1,59 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Purpose: Exploratory Data Analaysis for School Shooting Data
+# Author: Yun Chu
+# Date: 25 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
+# Pre-requisites: 03-clean_data.R has been run
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
-
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+data <- read_csv("data/02-analysis_data/analysis_data.csv")
 
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+# Dimensions of the dataset
+dim(data)
+
+# Column names
+colnames(data)
+
+# Check data structure
+str(data)
+
+# Summary of missing values
+colSums(is.na(data))
+
+# Summary for numerical columns
+summary(select_if(data, is.numeric))
+
+# Summary for categorical columns
+summary(select_if(data, is.character))
+
+# Incidents over Time
+ggplot(data, aes(x = date)) +
+  geom_histogram(binwidth = 30, fill = 'steelblue', color = 'black') +
+  labs(title = "Number of Incidents Over Time", x = "Date", y = "Count") +
+  theme_minimal()
+
+# Incidents by State
+# Count incidents by state
+state_counts <- data %>%
+  group_by(state) %>%
+  summarise(count = n()) %>%
+  arrange(desc(count))
+
+# Plot
+ggplot(state_counts, aes(x = reorder(state, count), y = count)) +
[…]

### 2024-11-26 scripts/06-model_data.R

@@ -1,36 +1,81 @@
 #### Preamble ####
 # Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Author: Yun Chu
+# Date: 25 November 2024
+# Contact: yun.chu@mail.utoronto.ca
+# License: CC BY-NC-SA 4.0
+# Pre-requisites: 03-clean_data.R has been run
+# Any other information needed? None
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
+library(dplyr)
+library(randomForest)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+data <- read_csv("data/02-analysis_data/analysis_data.csv")
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
 
+# Add year column and aggregate data by state and year
+data <- data %>%
+  group_by(state, year) %>%
+  summarise(
+    killed = sum(killed, na.rm = TRUE),
+    injured = sum(injured, na.rm = TRUE),
+    casualties = sum(casualties, na.rm = TRUE)
+  ) %>%
+  ungroup()
+
+# Prepare data for modeling
+train_data <- data %>% filter(year <= 2024)
+
+# Convert state to a factor (categorical variable)
+train_data <- train_data %>%
+  mutate(state = as.factor(state))
+
+# Define features and target variable
+X <- train_data %>% select(year, state)
+y <- train_data$casualties
+
+# Split data into training and test sets
+set.seed(42)
+train_indices <- sample(nrow(train_data), 0.8 * nrow(train_data))
+X_train <- X[train_indices, ]
+y_train <- y[train_indices]
+X_test <- X[-train_indices, ]
+y_test <- y[-train_indices]
+
[…]

### 2024-11-26 scripts/07-replications.R

@@ -1,4 +1,4 @@
-#### Preamble ####
+ #### Preamble ####
 # Purpose: Replicated graphs from... [...UPDATE THIS...]
 # Author: Rohan Alexander [...UPDATE THIS...]
 # Date: 11 February 2023 [...UPDATE THIS...]

### 2024-11-28 paper/paper.qmd
update model section

@@
 library(maps)
-
+library(knitr)
+library(randomForest)
+library(here)
 #### Read data ####
@@
 - maps
+- knitr
 
@@
 
-$ Causalities = Killed + Injured $
+In this dataset, there are three variables that have relationship: $ Causalities = Killed + Injured $.
+
+@tbl-casualties_summary_statistics summarizes the mean, median and standard deviation of casualties from school shooting events in US from 1999 to today.
+The standard deviation of 3.72 indicates that while most incidents have casualties close to the mean, there is a wide range of variability, with some incidents having significantly higher number of casualties.
+
+```{r}
+#| label: tbl-casualties_summary_statistics
+#| fig-cap: "Summary Statistics for Casualties"
+#| echo: false
+
+# Ensure the data contains the 'casualties' column
+# Calculate the mean, median, and standard deviation
+casualties_summary <- data %>%
+  summarise(
+    Mean = mean(casualties, na.rm = TRUE),
+    Median = median(casualties, na.rm = TRUE),
+    `Standard Deviation` = sd(casualties, na.rm = TRUE)
+  )
+
+# Display the summary in a beautiful table
+kable(casualties_summary, caption = "Summary Statistics for Casualties") %>%
+  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
+```
 
@@
 The outcome variable for this analysis is number of causalities for each state in 2025.
-
-@fig-Number_of_Incidents_Over_Time shows the number of school shooting incidents in US since 1999. The number of school shooting incidents almost doubled since 2020 compared with previous years.
-
-```{r}
-#| label: fig-Number_of_Incidents_Over_Time
-#| fig-cap: "Number of US School Shooting Incidents Over Time"
-#| echo: false
-
-ggplot(data, aes(x = date)) +
-  geom_histogram(binwidth = 60, fill = '#3FA0FF', color = 'white', alpha = 0.8) +
-  labs(
-    title = "Number of US School Shooting Incidents Over Time",
-    subtitle = "Visualizing school shooting trends in the US",
-    x = "Date",
-    y = "Incident Count"
-  ) +
-  theme_minimal(base_size = 14) +
-  theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 15)),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 10)),
-    axis.title.x = element_text(margin = margin(t = 10)),
-    axis.title.y = element_text(margin = margin(r = 10)),
-    panel.grid.major = element_line(color = "#E5E5E5"),
-    panel.grid.minor = element_blank()
-  )
-
-```
 
@@
 
-# Ensure state names are in lowercase to match map_data
+# Ensure state names in state_counts are lowercase to match map_data
 state_counts <- state_counts %>%
@@
 
-# Replace NA values in `total_casualties` with 0 (for states with no recorded casualties)
+# Replace missing values in `total_casualties` with 0 for states with no data
 map_data$total_casualties[is.na(map_data$total_casualties)] <- 0
 
+# Plot the map
 ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
@@
   scale_fill_gradient(
-    low = "#E6F7FF",
-    high = "#0072B2",
+    low = "#E6F7FF",  # Light blue
+    high = "#0072B2",  # Dark blue
     name = "Total Casualties",
-    breaks = c(0, 50, 100, 150, 200)  # Adjust the breaks based on your data
+    breaks = c(0, 50, 100, 150, 200)  # Adjust breaks based on your data
   ) +
@@
 
-Add graphs, tables and text.
+Predictor variables in this analysis is state and year.
 
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
+ - State represents the state where the schools are located.
+
+ - Year represents the shooting year.
+
+@fig-Number_of_Incidents_Over_Time shows the number of school shooting incidents in US since 1999. The number of school shooting incidents almost doubled since 2020 compared with previous years.
+
+```{r}
+#| label: fig-Number_of_Incidents_Over_Time
+#| fig-cap: "Number of US School Shooting Incidents Over Time"
+#| echo: false
+#| eval: false
+
+ggplot(data, aes(x = date)) +
+  geom_histogram(binwidth = 60, fill = '#3FA0FF', color = 'white', alpha = 0.8) +
+  labs(
+    title = "Number of US School Shooting Incidents Over Time",
+    subtitle = "Visualizing school shooting trends in the US",
+    x = "Date",
+    y = "Incident Count"
+  ) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 15)),
+    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 10)),
+    axis.title.x = element_text(margin = margin(t = 10)),
+    axis.title.y = element_text(margin = margin(r = 10)),
+    panel.grid.major = element_line(color = "#E5E5E5"),
+    panel.grid.minor = element_blank()
+  )
+
+```
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
+## Model Set-Up
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+To model the total number of casualties resulting from school shooting incidents, a **Random Forest regression model** is employed.
+The model is designed to predict casualties based on state-level and temporal features derived from the dataset.
+The Random Forest algorithm is chosen for its robustness to non-linear relationships, its ability to handle interactions among predictors, and its ability to integrate categorical variables without creating dummy variables.
 
-## Model set-up
+The mathematical representation of the model is: $C = f(X_1, X_2, ..., X_p) + \epsilon$
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+where:
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+- $C$: The total number of casualties (sum of fatalities and injuries) for a given year and state.  
+- $f(\cdot)$: The function estimated by the Random Forest model, representing the aggregate prediction from all decision trees.  
+- $X_1, X_2, ..., X_p$: Predictors, including:  
+  - $X_1$: Year (numeric), capturing temporal trends.  
+  - $X_2$: State (categorical), capturing regional effects.  
+- $\epsilon$: The error term, representing unobserved factors affecting the number of casualties.
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+## Model Justification
+- **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
+- **State** is treated as a categorical variable to capture location-specific effects, such as differences in legislation, policing, or socioeconomic factors.
 
-### Model justification
+The choice of these features aligns with the data section, ensuring that variables with potential predictive value are incorporated without overfitting or introducing unnecessary complexity.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+### Assumptions and Limitations
+- **Assumptions**:
+  1. Casualties are conditionally independent given the predictors.
+  2. The relationship between predictors and casualties can be approximated by the ensemble of decision trees.
+  3. Data is representative of underlying trends and free of major sampling biases.
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+- **Limitations**:
+  1. The model may not generalize well to future data if underlying trends shift dramatically (e.g., policy changes or societal events).
+  2. Random Forests lack explicit interpretability compared to simpler models.
+
+### Software Implementation
+The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.
+
+### Model Validation
+- **Train/Test Splits**:
+  The data is split into training (80%) and test (20%) sets. The model is trained on the former and validated on the latter to ensure generalizability.
+
+- **Evaluation Metrics**:
+  - **Root Mean Squared Error (RMSE)** and **Mean Absolute Error (MAE)** are calculated to assess prediction accuracy.
+  - Final metrics:
+    - RMSE: 1.491141 casualties
+    - MAE: 1.096164 casualties
+
+### Alternative Models Considered
+- **Linear Regression**:
+  - Strengths: Simplicity and interpretability.
+  - Weaknesses: Inadequate for capturing non-linear relationships and interactions present in the data.
+
+- **Gradient Boosting Machines (GBM)**:
+  - Strengths: Often more accurate due to its iterative learning.
+  - Weaknesses: Higher risk of overfitting, more computationally expensive and less intepretibility.
+
+The Random Forest model was chosen over these alternatives for its balance of flexibility, robustness, and interpretability of feature importance.
 
@@
 #| eval: true
+#| label: fig-predicted_shooting_visualization
+#| tbl-cap: "Predicted 2025 US School Shootings Numbers"
 #| warning: false
-#| message: false
 
-library(rstanarm)
+#### Step 1: Load the Data and Model ####
+# Load cleaned data
+clean_data <- read.csv("../data/02-analysis_data/no_outlier_data.csv")
 
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
+# Ensure 'state' is a factor
+clean_data$state <- as.factor(clean_data$state)
+
+# Load the Random Forest model
+rf_model_clean <- readRDS(file = here::here("models/rf_model.rds"))
+
+#### Step 2: Prepare Data for Predictions ####
+# Year(s) for prediction
+years <- 2025
+
+# Create a data frame with unique states and years
+prediction_data <- expand.grid(
+  year = years,
+  state = unique(clean_data$state)  # Use unique states
+)
+
+# Convert 'state' to a factor and match levels with training data
+prediction_data$state <- factor(prediction_data$state, levels = levels(clean_data$state))
+
+# Ensure no missing levels
+if (any(is.na(prediction_data$state))) {
+  stop("State levels in prediction_data do not match clean_data.")
+}
+
+# Make predictions
+predictions <- predict(rf_model_clean, newdata = prediction_data)
+
+# Add predictions to the data frame
+prediction_data$predicted_casualties <- predictions
+
+#### Step 3: Prepare Map Data ####
+# Get US map data
+us_map <- map_data("state")
+
+# Convert state names in `us_map` to lowercase
+us_map$region <- tolower(us_map$region)
+
+# Convert state in `prediction_data` to lowercase for compatibility
+prediction_data$state <- tolower(as.character(prediction_data$state))
+
+# Merge map data with predictions
+map_data <- us_map %>%
+  left_join(prediction_data, by = c("region" = "state"))
+
+# Fill missing predictions with 0 (optional, avoids grey states)
+map_data$predicted_casualties[is.na(map_data$predicted_casualties)] <- 0
+
+#### Step 4: Plot the Map ####
+ggplot(map_data, aes(long, lat, group = group, fill = predicted_casualties)) +
+  geom_polygon(color = "white") +
+  scale_fill_gradient(
+    low = "#E6F7FF",  # Light blue
+    high = "#0072B2", # Dark blue
+    name = "Casualties"
+  ) +
+  labs(
+    title = "Predicted US School Shooting Casualties by State (2025)",
+    subtitle = "Number of casualties visualized geographically",
+    x = NULL,
+    y = NULL
+  ) +
+  coord_fixed(1.3) +
+  theme_minimal(base_size = 14) +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
+    plot.subtitle = element_text(hjust = 0.5, size = 10),
+    legend.position = "right",
+    axis.text = element_blank(),
+    axis.ticks = element_blank(),
+    panel.grid = element_blank()
+  )
+
 ```

### 2024-11-28 scripts/06-model_data.R

@@ -39,44 +39,84 @@ train_data <- train_data %>%
 X <- train_data %>% select(year, state)
 y <- train_data$casualties
 
-# Split data into training and test sets
+#### Cross-Validation for Residual Calculation ####
+
+# Define cross-validation folds
+library(caret)
 set.seed(42)
-train_indices <- sample(nrow(train_data), 0.8 * nrow(train_data))
-X_train <- X[train_indices, ]
-y_train <- y[train_indices]
-X_test <- X[-train_indices, ]
-y_test <- y[-train_indices]
+folds <- createFolds(y, k = 5, list = TRUE)
 
-# Train a Random Forest model
-rf_model <- randomForest(X_train, y_train, ntree = 300)
+# Calculate residuals for each fold
+residuals_all <- numeric()
+for (fold in folds) {
+  train_idx <- setdiff(seq_len(nrow(train_data)), fold)
+  test_idx <- fold
+  
+  X_train_cv <- X[train_idx, ]
+  y_train_cv <- y[train_idx]
+  X_test_cv <- X[test_idx, ]
+  y_test_cv <- y[test_idx]
+  
+  # Train Random Forest on CV fold
+  rf_model_cv <- randomForest(X_train_cv, y_train_cv, ntree = 300)
+  
+  # Predict and calculate residuals
+  y_pred_cv <- predict(rf_model_cv, newdata = X_test_cv)
+  residuals_cv <- y_test_cv - y_pred_cv
+  residuals_all[test_idx] <- residuals_cv
+}
 
-# Evaluate the model on the test set
-y_pred <- predict(rf_model, newdata = X_test)
-mae <- mean(abs(y_test - y_pred))
-rmse <- sqrt(mean((y_test - y_pred)^2))
+# Add residuals to the training data
+train_data$residuals <- residuals_all
 
-# Print evaluation metrics
-cat("Model Evaluation:\n")
-cat("Mean Absolute Error (MAE):", mae, "\n")
-cat("Root Mean Squared Error (RMSE):", rmse, "\n")
+#### Identify and Remove Outliers ####
+# Set a threshold for residuals (e.g., absolute residual > 10)
+outlier_threshold <- 10
+clean_data <- train_data %>% filter(abs(residuals) <= outlier_threshold)
 
-# Prepare future data for 2025 predictions
-future_data <- data.frame(
-  year = 2025,
-  state = unique(train_data$state)  # Predict for all states
-)
[…]

### 2024-11-28 paper/paper.qmd
update intro, discussion and result section

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "School shootings in the United States pose a critical challenge with far-reaching impacts.
+This paper employs a Random Forest regression model to predict state-level casualties for 2025 using historical data from 1999 to the present.
+The findings reveal a positively skewed distribution of casualties, with states like Georgia, California, and Pennsylvania projected to face the highest numbers.
+These insights highlight the need for targeted interventions, including enhanced school safety measures and stricter gun control policies.
+By providing actionable predictions, this study aims to inform policies that reduce school shooting casualties and protect vulnerable populations."
 format: pdf
-number-sections: true
+toc: true
 bibliography: references.bib
@@
 library(here)
+library(kableExtra)
 #### Read data ####
@@
 
-# Introduction
+# Introduction {#sec-intro}
 
-Overview paragraph
+School shootings remain a grave concern in the United States, with devastating consequences for students, educators, and communities.
+Despite the growing public discourse and policy debates surrounding school safety, little progress has been made in effectively predicting and preventing these tragic incidents.
+Understanding the patterns and predictors of school shootings is essential for designing evidence-based interventions to protect students and reduce casualties.
+This paper contributes to this critical issue by leveraging data-driven methodologies to analyze and predict school shooting casualties, with the aim of informing policy and prevention efforts.
 
-Estimand paragraph
+The estimand in this paper is the number of casualties from school shootings across U.S.
+states for the year 2025.
+Using historical data from The Washington Post [@washingtonpost_school_shootings], which provides a comprehensive record of school shootings, a Random Forest regression model was developed to predict casualties based on state and temporal features.
+The analysis includes summary statistics, spatial visualizations, and predictive modeling, providing actionable insights for policymakers.
+This study fills a critical gap in understanding the state-level distribution of school shooting casualties and offers data-driven recommendations for targeted interventions.
 
-Results paragraph
+A key finding of this study is the uneven distribution of casualties across states, with some states, such as Georgia, California, and Pennsylvania, predicted to experience significantly higher casualties in 2025.
+This highlights the importance of state-specific policies and resource allocation.
+The analysis also reveals that the distribution of casualties is positively skewed, with most incidents involving relatively few casualties, but a small number of high-casualty events disproportionately contributing to the overall impact.
+This information underscore the need for targeted strategies to mitigate the effects of high-casualty incidents.
 
-Why it matters paragraph
+The importance of this study lies in its ability to provide a data-driven foundation for policy and prevention efforts.
+By identifying high-risk states and understanding the characteristics of high-casualty incidents, this paper aims to guide policymakers in allocating resources effectively and designing interventions that address the underlying factors contributing to school shootings.
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
-
-Paragraph of Estimand!
+The structure of this paper is as follows: @sec-data discusses the data, including its sources, measurements, and key variables; @sec-model outlines the model used for prediction, including its justification and validation; @sec-result presents the results, including summary statistics, spatial visualizations, and predictions for 2025; and @sec-discussion concludes with a discussion of the findings, limitations, and recommendations for future research and policy.
 
@@
 
-## Overview
+## Overview {#sec-data-overview}
 
@@
 - randomForest [@randomForest2024]
-- ggplot2
-- maps
-- knitr
+- ggplot2 [@ggplot2]
+- maps [@maps]
+- knitr [@knitr]
+- here [@here]
+- kableExtra [@kableExtra]
 
-## Measurement
+## Measurement {#sec-data-measurement}
 
@@
 
-## Summary Statistics & Relationship Between Variables
+## Summary Statistics & Relationship Between Variables {#sec-data-sumsta}
 
@@
 
-## Outcome variables
+## Outcome variables {#sec-data-outvariable}
 
@@
 
-## Predictor variables
+## Predictor variables {#sec-predic-variable}
 
@@
 
-# Model
+# Model {#sec-model}
 
-## Model Set-Up
+## Model Set-Up {#sec-model-setup}
 
@@
 
-## Model Justification
+## Model Justification {#sec-model-justification}
 - **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
@@
 
+Number of trees is selected to be 300 to balance accuracy and prevent overfitting.
+
 The choice of these features aligns with the data section, ensuring that variables with potential predictive value are incorporated without overfitting or introducing unnecessary complexity.
 
-### Assumptions and Limitations
+### Assumptions and Limitations {#sec-model-assumplim}
 - **Assumptions**:
@@
 
-### Software Implementation
+### Software Implementation {#sec-model-softimp}
 The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.
 
-### Model Validation
+### Model Validation {#sec-model-modelvalidation}
 - **Train/Test Splits**:
@@
 
-### Alternative Models Considered
+### Alternative Models Considered {#sec-model-altmodel}
 - **Linear Regression**:
@@
 
-# Results
+# Results {#sec-result}
 
-Our results are summarized in @tbl-modelresults.
+Since random forest is non-parametric, there is not parameter estimates. ????
+
+## Summary Statistics {#sec-result-summstat}
+
+@tbl-summary_statistics_table shows the summary statistics for predicted 2025 school shooting casualties.
+On average, each school shooting incident results in approximately 1.59 casualties.
+At least half of the incidents involve only one casualty.
+The standard deviation of 1.91 indicates that while many incidents cluster around the mean of 1.59, there is considerable spread, with some incidents resulting in significantly higher casualties.
 
@@
 #| eval: true
+#| label: tbl-summary_statistics_table
+#| tbl-cap: "Summary Statistics for 2025 US School Shooting Casualties"
+#| warning: false
+
+# Load cleaned data
+clean_data <- read.csv("../data/02-analysis_data/no_outlier_data.csv")
+
+# Calculate summary statistics for the 'casualties' column
+summary_stats <- clean_data %>%
+  summarise(
+    Mean = mean(casualties, na.rm = TRUE),
+    Median = median(casualties, na.rm = TRUE),
+    `Standard Deviation` = sd(casualties, na.rm = TRUE)
+  )
+
+# Display the summary statistics as a table
+summary_stats %>%
+  kable(
+    format = "latex", 
+    booktabs = TRUE, 
+  ) %>%
+  kable_styling(
+    latex_options = c("striped"), 
+    full_width = FALSE
+  )
+
+```
+
+## Predicted Casualties by State {#sec-result-predicted}
+
+Using a Random Forest model, predictions for total casualties in each state for the year 2025 were generated.
+The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization.
+States such as California and Texas are predicted to experience the highest casualties, reflecting historical trends.
+```{r}
+#| echo: false
+#| eval: true
 #| label: fig-predicted_shooting_visualization
-#| tbl-cap: "Predicted 2025 US School Shootings Numbers"
+#| fig-cap: "Predicted 2025 US School Shootings Numbers"
 #| warning: false
@@
 
+The predicted casualties for each state are summarized in @tbl-predicted_shooting_table.
+The table provides a clear and concise overview, sorted in descending order of predicted casualties to emphasize the most affected states.
+
 ```{r}
 #| echo: false
-#| eval: false
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
+#| eval: true
+#| label: tbl-predicted_shooting_table
+#| tbl-cap: "Predicted School Shooting Casualties by State for 2025"
 #| warning: false
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
+# Prepare the predictions table without the year column
+prediction_table <- prediction_data %>%
+  select(-year) %>%  # Remove the 'year' column
+  arrange(desc(predicted_casualties)) %>%
+  rename(State = state, "Predicted Casualties" = predicted_casualties)
+
+# Display the table using kable
+prediction_table %>%
+  kable(format = "latex", booktabs = TRUE) %>%
+  kable_styling(latex_options = c("striped"))  # Removed "hold_position"
 ```
 
-# Discussion
+# Discussion {#sec-discussion}
 
-## First discussion point {#sec-first-point}
+## Actions to mitigate the impact of high-casualty incidents {#sec-discussion-first-point}
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The distribution of school shooting casualties reveals significant insights into the nature of these incidents.
+The mean number of casualties is higher than the median, indicating a positively skewed distribution.
+This suggests that while most school shooting incidents result in relatively few casualties, there are a small number of events that result in significantly higher numbers of casualties.
+These high-casualty incidents have a disproportionate impact and should be a focal point for policymakers.
+Understanding the causes and contexts of such incidents, such as the presence of high-density schools, specific socio-economic factors, or insufficient security measures, could help in designing targeted interventions.
+For instance, schools in high-risk areas could benefit from increased security personnel, improved mental health resources for students, and stricter firearm regulations.
+Government must prioritize strategies aimed at mitigating the impact of these high-casualty incidents to protect vulnerable populations.
 
-## Second discussion point
+## Targeted Interventions for High-Risk States {#sec-discussion-sec-point}
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+The state-level predictions of casualties, as visualized in the map and table, provide actionable insights for policymakers at both the state and federal levels.
+States such as Georgia, California, and Pennsylvania are projected to experience higher numbers of casualties in 2025.
+This highlights the need for state-specific policies and resource allocation.
+For example, states with higher predicted casualties could implement tailored programs such as increasing mental health support in schools, conducting safety drills, and investing in school infrastructure to enhance safety measures.
+Additionally, identifying regions with consistently higher casualties can help governments allocate resources more effectively and conduct targeted investigations to address root causes.
 
-## Third discussion point
+## Policy Recommendations to Reduce Casualties {#sec-discussion-third-point}
 
-## Weaknesses and next steps
+The map and table of predicted casualties highlight opportunities for governments to take preventative action.
+One key recommendation is to introduce universal background checks and stricter gun control measures in states with high predicted casualties.
+Furthermore, funding programs that promote community engagement and early intervention in at-risk populations could help reduce the likelihood of school shootings.
+Schools can also be equipped with better surveillance technology and emergency response systems to minimize casualties during incidents.
+At a broader level, creating national frameworks for school safety policies can ensure consistency and accountability in protecting students and staff across all states.
 
-Weaknesses and next steps should also be included.
+## Weaknesses and next steps {#sec-discussion-weaknessnext}
+
+While the Random Forest model provided reasonable predictions, it exhibited some limitations.
+The negative percentage of variance explained during initial iterations indicated that the model struggled to generalize with the available predictors.
+Although this was addressed by refining the data and removing outliers, the inclusion of only year and state as predictors limits the model's explanatory power.
+Additionally, the dataset's small size and imbalance across states might have impacted the model's ability to identify robust patterns.
+
+The findings from this study underscore the importance of integrating predictive modeling with proactive policy-making.
+However, future work should focus on enhancing the dataset by incorporating additional predictors such as gun ownership rates, local crime statistics, and school-specific characteristics.
+
+Additionally, examining the effectiveness of existing policies in high-risk areas could provide valuable insights into which interventions are most impactful.
+
+Finally, collaborative efforts between federal agencies, state governments, and local communities will be essential to addressing the systemic issues underlying school shootings and reducing the number of casualties in future incidents.
 

### 2024-11-28 paper/paper.qmd
update paper

@@
 #| echo: false
-#| eval: false
+#| eval: true
 
@@
   theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 15)),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 10)),
-    axis.title.x = element_text(margin = margin(t = 10)),
-    axis.title.y = element_text(margin = margin(r = 10)),
+    plot.title = element_text(
+      hjust = 0.5, 
+      face = "bold", 
+      size = 13
+    ),
+    plot.subtitle = element_text(
+      hjust = 0.5, 
+      size = 10
+    ),
+    axis.title.x = element_text(),
+    axis.title.y = element_text(),
     panel.grid.major = element_line(color = "#E5E5E5"),

### 2024-12-03 paper/paper.qmd
paper & model update

--- 2024-11-28 paper/paper.qmd
@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "School shootings in the United States pose a critical challenge with far-reaching impacts.
+This paper employs a Random Forest regression model to predict state-level casualties for 2025 using historical data from 1999 to the present.
+The findings reveal a positively skewed distribution of casualties, with states like Georgia, California, and Pennsylvania projected to face the highest numbers.
+These insights highlight the need for targeted interventions, including enhanced school safety measures and stricter gun control policies.
+By providing actionable predictions, this study aims to inform policies that reduce school shooting casualties and protect vulnerable populations."
 format: pdf
+toc: true
 number-sections: true
@@
 library(knitr)
-library(randomForest)
+library(modelsummary)
 library(here)
+library(kableExtra)
+library(arrow)
 #### Read data ####
-data <- read_csv("../data/02-analysis_data/analysis_data.csv")
+data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
+data$ifcasualty <- ifelse(data$casualties> 0, 1, 0)
+# Create a new column with simplified categories
+data$shooting_category <- ifelse(
+  data$shooting_type %in% c("indiscriminate", "targeted"),  
+  "indiscriminate or targeted",                              
+  "other"
+)
+
 ```
 
-# Introduction
+# Introduction {#sec-intro}
 
-Overview paragraph
+School shootings remain a grave concern in the United States, with devastating consequences for students, educators, and communities.
+Despite the growing public discourse and policy debates surrounding school safety, little progress has been made in effectively predicting and preventing these tragic incidents (Jonson, 2017; RAND Corporation, 2015).
+Understanding the patterns and predictors of school shootings is essential for designing evidence-based interventions to protect students and reduce casualties.
+This paper contributes to this critical issue by leveraging data-driven methodologies to analyze and predict the existence of school shooting casualties, with the aim of informing policy and prevention efforts.
 
-Estimand paragraph
+The estimand in this paper is to classify whether casualty exists in a school shooting incident in US.
+Using historical data from The Washington Post [@washingtonpost_school_shootings], which provides a comprehensive record of school shootings, a Bayesian Logistic Regression model was developed to predict the existence of casualty based on shooting type and latitude of the school.
+The analysis includes summary statistics, spatial visualizations, and predictive modeling, providing actionable insights for policymakers.
+This study fills a critical gap in understanding the probability of school shooting casualties and offers data-driven recommendations for targeted interventions.
 
-Results paragraph
+A key finding of this study is the uneven distribution of casualties across states, with some states, such as Georgia, California, and Pennsylvania, predicted to experience significantly higher casualties in 2025.
+This highlights the importance of state-specific policies and resource allocation.
+The analysis also reveals that the distribution of casualties is positively skewed, with most incidents involving relatively few casualties, but a small number of high-casualty events disproportionately contributing to the overall impact.
+This information underscore the need for targeted strategies to mitigate the effects of high-casualty incidents.
 
-Why it matters paragraph
+The importance of this study lies in its ability to provide a data-driven foundation for policy and prevention efforts.
+By identifying high-risk states and understanding the characteristics of high-casualty incidents, this paper aims to guide policymakers in allocating resources effectively and designing interventions that address the underlying factors contributing to school shootings.
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
-
-Paragraph of Estimand!
+The structure of this paper is as follows: @sec-data discusses the data, including its sources, measurements, and key variables; @sec-model outlines the model used for prediction, including its justification, model validation is included in @sec-appendix; @sec-result presents the results, including summary statistics and spatial visualizations; and @sec-discussion concludes with a discussion of the findings, limitations, and recommendations for future research and policy.
 
@@
 Additionally, shootings at colleges and universities, which involve young adults rather than children, were not included in the analysis [@washingtonpost_school_shootings].
-These entries cover 50 variables that provide information about the schools, locations, date and time of shooting, shooters details, number of killed and injured, and the relationship of the shooter to school, the weapon type and source.
+These entries cover 50 variables that provide information about the schools and its students, locations, date and time of shooting, shooters details, number of killed and injured, and the relationship of the shooter to school, the weapon type and source.
 
 As the federal government does not consistently track school shootings, this dataset from *The Washington Post* fills a critical gap.
-It was carefully assembled using information from diverse sources, including news articles, open-source databases, law enforcement reports, and direct inquiries to schools and police departments.
+It was carefully assembled using information from diverse sources, including news articles, open-source databases, law enforcement reports, and direct inquiries to schools and police departments [@washingtonpost_school_shootings].
 Although sources like FBI crime reports and local school incident logs were reviewed, they lack the detail and comprehensive coverage of this dataset.
@@
 
-We use the statistical programming language R [@citeR] to download, clean, analyze and model the US School Shooting Data.
+The statistical programming language R [@citeR] is used to download, clean, analyze and model the US School Shooting Data.
 The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] .
@@
 - testthat [@testthat2024]
-- randomForest [@randomForest2024]
-- ggplot2
-- maps
-- knitr
+- modelsummary [@modelsummary]
+- ggplot2 [@ggplot2]
+- maps [@maps]
+- knitr [@knitr]
+- here [@here]
+- kableExtra [@kableExtra]
+- rstanarm [@rstanarm]
 
-## Measurement
+## Measurement {#sec-data-measurement}
 
@@
 
-## Summary Statistics & Relationship Between Variables
+## Summary Statistics & Relationship Between Variables {#sec-data-sumsta}
 
-In this dataset, there are three variables that have relationship: $ Causalities = Killed + Injured $.
+In this dataset, there are three variables that have relationship:
+$causalities = killed + knjured$.
 
-@tbl-casualties_summary_statistics summarizes the mean, median and standard deviation of casualties from school shooting events in US from 1999 to today.
+@tbl-casualties_summary_statistics summarizes the mean, median and standard deviation of casualties from school shooting events in US from 1999 till now.
 The standard deviation of 3.72 indicates that while most incidents have casualties close to the mean, there is a wide range of variability, with some incidents having significantly higher number of casualties.
@@
 #| label: tbl-casualties_summary_statistics
-#| fig-cap: "Summary Statistics for Casualties"
+#| tbl-cap: "Summary Statistics for Casualties"
 #| echo: false
@@
 # Ensure the data contains the 'casualties' column
-# Calculate the mean, median, and standard deviation
+# Calculate the mean, median, and standard deviation with two decimal places
 casualties_summary <- data %>%
   summarise(
-    Mean = mean(casualties, na.rm = TRUE),
-    Median = median(casualties, na.rm = TRUE),
-    `Standard Deviation` = sd(casualties, na.rm = TRUE)
+    Mean = round(mean(casualties, na.rm = TRUE), 2),
+    Median = round(median(casualties, na.rm = TRUE), 2),
+    `Standard Deviation` = round(sd(casualties, na.rm = TRUE), 2)
   )
 
-# Display the summary in a beautiful table
+# Display the summary in a styled table
 kable(casualties_summary, caption = "Summary Statistics for Casualties") %>%
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
+  kableExtra::kable_styling(
+    bootstrap_options = c("striped", "hover", "condensed"),
+    full_width = FALSE,
+    position = "center"
+  )
+```
+@fig-casualty-by-shooting-type illustrates the distribution of casualties across different shooting types.
+The "Targeted" and "Indiscriminate" categories exhibit significantly higher casualty counts compared to other types.
+As a result, these two types are combined into one class, while the remaining types form a separate class.
+
+```{r}
+#| label: fig-casualty-by-shooting-type
+#| fig-cap: "Distribution of Casualty By Shooting Type"
+#| echo: false
+#| eval: true
+#| 
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+
+# Summarize the total casualties for all shooting types
+casualties_by_type <- data %>%
+  group_by(shooting_type) %>%
+  summarise(total_casualties = sum(casualties, na.rm = TRUE)) %>%
+  arrange(desc(total_casualties))  # Arrange by descending number of casualties
+
+# Plot the total number of casualties by shooting type
+ggplot(casualties_by_type, aes(x = reorder(shooting_type, total_casualties), y = total_casualties, fill = shooting_type)) +
+  geom_bar(stat = "identity", color = "white") +
+  scale_fill_brewer(palette = "Set3") +
+  labs(
+    title = "Number of Casualties by Shooting Type",
+    x = "Shooting Type",
+    y = "Total Number of Casualties",
+    fill = "Shooting Type"
+  ) +
+  theme_minimal() +
+  theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
+    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for readability
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    legend.position = "none"  # Hide legend as labels are self-explanatory
+  )
 ```
 
-## Outcome variables
+## Outcome variables {#sec-data-outvariable}
 
-The outcome variable for this analysis is number of causalities for each state in 2025.
+The outcome variable for this analysis is causalities for each state in 2025. In the model, we classify the existence of casualties.
+
+@fig-casualtydist illustrates the distribution of casualties.
+The data is positively skewed, with the majority of cases involving fewer than 7 casualties, though there are a few school shooting incidents with exceptionally high casualty counts.
+
+```{r}
+#| label: fig-casualtydist
+#| echo: false
+#| fig-cap: Distribution of Casualties
+# Causaulity Distribution
+ggplot(data, aes(x = casualties)) +
+  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
+  labs(x = "Number of Casualties", y = "Frequency") +
+  theme_minimal()
+
+```
 
@@
 
-# Aggregate data to calculate total casualties by state
 state_counts <- data %>%
@@
 
-# Replace missing values in `total_casualties` with 0 for states with no data
+# Replace missing values in total_casualties with 0 for states with no data
 map_data$total_casualties[is.na(map_data$total_casualties)] <- 0
@@
 
-## Predictor variables
+## Predictor variables {#sec-predic-variable}
 
-Predictor variables in this analysis is state and year.
+Predictor variables in this analysis is Shooting Category and Latitude.
 
- - State represents the state where the schools are located.
+ - Shooting Category is a binary variable derived from Shooting Type with "Targeted" and "Indiscriminate" as one type while the rest is classified as the second type.
+Shooting Type represents the type of shootings with 5 categories: accidental, hostage suicide, indiscriminate, public suicide, targeted, and unclear.
+There are also some rows with a combination of these 5 types, for example, "accidental or targeted" and "targeted and indiscriminate".
 
- - Year represents the shooting year.
+ - Latitude represents the latitude of the school.
 
-@fig-Number_of_Incidents_Over_Time shows the number of school shooting incidents in US since 1999. The number of school shooting incidents almost doubled since 2020 compared with previous years.
+@fig-shootingtypedist presents a pie chart depicting the proportions of various shooting types. Targeted shootings account for 56% of all school shooting incidents.
 
 ```{r}
-#| label: fig-Number_of_Incidents_Over_Time
-#| fig-cap: "Number of US School Shooting Incidents Over Time"
+#| label: fig-shootingtypedist
+#| fig-cap: "Shooting Type Distribution" 
 #| echo: false
-#| eval: false
+#| eval: true
 
-ggplot(data, aes(x = date)) +
-  geom_histogram(binwidth = 60, fill = '#3FA0FF', color = 'white', alpha = 0.8) +
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+library(ggrepel)
+
+# Simulated dataset (replace this with your actual data)
+# Assuming `data` has a column `shooting_type`
+shooting_data <- data %>% 
+  count(shooting_type) %>% 
+  mutate(
+    proportion = n / sum(n),  # Calculate proportions
+    label = paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")")  # Labels
+  )
+
+# Calculate label positions and angles
+shooting_data <- shooting_data %>%
+  mutate(
+    y_position = cumsum(proportion) - 0.5 * proportion,  # Calculate mid-point of each segment
+    angle = 90 - (360 * (cumsum(proportion) - proportion / 2))  # Adjust angles for better label alignment
+  )
+
+# Plot Pie Chart
+ggplot(shooting_data, aes(x = "", y = proportion, fill = shooting_type)) +
+  geom_bar(width = 1, stat = "identity", color = "white") +
+  coord_polar(theta = "y") +
+  theme_void() +
+  scale_fill_brewer(palette = "Set3") +
   labs(
-    title = "Number of US School Shooting Incidents Over Time",
-    subtitle = "Visualizing school shooting trends in the US",
-    x = "Date",
-    y = "Incident Count"
+    title = "Distribution of Shooting Types",
+    fill = "Shooting Type"
   ) +
-  theme_minimal(base_size = 14) +
+  geom_text(
+    aes(
+      label = ifelse(proportion > 0.05, paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")"), ""),
+      y = y_position,
+      angle = ifelse(proportion < 0.1, angle, 0)  # Add dynamic rotation for small segments
+    ),
+    size = 3,
+    hjust = ifelse(shooting_data$proportion < 0.1, 1, 0.5),  # Align labels for smaller slices
+    nudge_x = 0.4  # Nudge outward for better alignment
+  ) +
   theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 15)),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 10)),
-    axis.title.x = element_text(margin = margin(t = 10)),
-    axis.title.y = element_text(margin = margin(r = 10)),
-    panel.grid.major = element_line(color = "#E5E5E5"),
-    panel.grid.minor = element_blank()
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    legend.title = element_text(size = 10),
+    legend.position = "right"
+  )
+```
+
+@fig-latitudedist shows the distribution of latitude values of schools in school shooting events, ranging from approximately 20 to 50 degrees.
+Most events occur between 30 and 40 degrees, with a peak near 35 degrees.
+There are few occurrences at higher latitudes (above 45 degrees).
+
+```{r}
+#| label: fig-latitudedist
+#| fig-cap: "Distribution of Latitude of Schools in School Shooting Events"
+#| echo: false
+#| eval: true
+#| warning: false
+
+# Load necessary libraries
+library(ggplot2)
+
+# Assuming `data` contains the `lat` column (latitude values)
+# Replace `data` with your actual dataset
+
+# Plot the distribution of latitude
+ggplot(data, aes(x = lat)) +
+  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
+  theme_minimal() +
+  labs(
+    title = "Distribution of Latitude",
+    x = "Latitude",
+    y = "Frequency"
+  ) +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12)
   )
@@
 
-# Model
+# Model {#sec-model}
 
-## Model Set-Up
+## Model Set-Up {#sec-model-setup}
 
-To model the total number of casualties resulting from school shooting incidents, a **Random Forest regression model** is employed.
-The model is designed to predict casualties based on state-level and temporal features derived from the dataset.
-The Random Forest algorithm is chosen for its robustness to non-linear relationships, its ability to handle interactions among predictors, and its ability to integrate categorical variables without creating dummy variables.
+The analysis employs a Bayesian Logistic Regression Model to predict whether a school shooting incident results in casualties.
+The binary response variable (`ifcasualty`) is defined as 1 if there are any casualties (injuries or fatalities) and 0 otherwise.
+The predictors include the **latitude** of the school where the incident occurs and a simplified categorical variable, **shooting category**, which distinguishes between "indiscriminate" or "targeted" shootings and "other" types.
 
-The mathematical representation of the model is: $C = f(X_1, X_2, ..., X_p) + \epsilon$
+### Mathematical Formulation
+
+The model estimates the probability of a casualty using the following formulation:
+
+$$
+P(C = 1 \mid X) = \frac{\exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2)}{1 + \exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2)}
+$$
+$$
+\beta_0 \sim \text{Normal}(0, 5)
+$$
+$$
+\beta_1, \beta_2 \sim \text{Normal}(0, 2.5)
+$$
 
@@
 
-- $C$: The total number of casualties (sum of fatalities and injuries) for a given year and state.  
-- $f(\cdot)$: The function estimated by the Random Forest model, representing the aggregate prediction from all decision trees.  
-- $X_1, X_2, ..., X_p$: Predictors, including:  
-  - $X_1$: Year (numeric), capturing temporal trends.  
-  - $X_2$: State (categorical), capturing regional effects.  
-- $\epsilon$: The error term, representing unobserved factors affecting the number of casualties.
+- $P(C = 1 \mid X)$: Probability of at least one casualty occurring in an incident.
+- $\beta_0$: Intercept term.
+- $\beta_1$: Coefficient for the shooting category, capturing the difference between "indiscriminate or targeted" and "other" types.
+- $\beta_2$: Coefficient for latitude, accounting for geographic trends.
+- $X_1$: Shooting category (0 = "indiscriminate or targeted", 1 = "other").
+- $X_2$: Latitude of the school where the incident occurs (continuous).
 
-## Model Justification
-- **Year** is included to account for temporal trends in school shootings, reflecting potential increases or decreases over time.
-- **State** is treated as a categorical variable to capture location-specific effects, such as differences in legislation, policing, or socioeconomic factors.
+The priors for the Bayesian logistic regression model are specified as follows:
 
-The choice of these features aligns with the data section, ensuring that variables with potential predictive value are incorporated without overfitting or introducing unnecessary complexity.
+- **Intercept ($\beta_0$)**: Reflects uncertainty about the baseline log-odds of casualties.
+- **Coefficients ($\beta_1, \beta_2$)**: Weakly informative priors that regularize the model and prevent overfitting.
 
-### Assumptions and Limitations
-- **Assumptions**:
-  1. Casualties are conditionally independent given the predictors.
-  2. The relationship between predictors and casualties can be approximated by the ensemble of decision trees.
-  3. Data is representative of underlying trends and free of major sampling biases.
+These priors were chosen to represent plausible parameter values while maintaining model stability.
 
-- **Limitations**:
-  1. The model may not generalize well to future data if underlying trends shift dramatically (e.g., policy changes or societal events).
-  2. Random Forests lack explicit interpretability compared to simpler models.
+The Bayesian logistic regression model is implemented in **R** using the `rstanarm` package. The final model is saved as an RDS file for reproducibility and further analysis.
 
-### Software Implementation
-The model is implemented in **R**, using the randomForest [@randomforest] package for training and evaluation.
+### Model Justification {#sec-model-justification}
 
-### Model Validation
-- **Train/Test Splits**:
-  The data is split into training (80%) and test (20%) sets. The model is trained on the former and validated on the latter to ensure generalizability.
+The Bayesian logistic regression model is well-suited for this problem as it is specifically designed to predict binary outcomes, such as whether casualties occur, without requiring transformation of skewed response variables.
+Additionally, the Bayesian framework provides probabilistic predictions, allows for the incorporation of prior knowledge, and offers a robust quantification of uncertainty in parameter estimates.
 
-- **Evaluation Metrics**:
-  - **Root Mean Squared Error (RMSE)** and **Mean Absolute Error (MAE)** are calculated to assess prediction accuracy.
-  - Final metrics:
-    - RMSE: 1.491141 casualties
-    - MAE: 1.096164 casualties
+The shooting types "Targeted" and "Indiscriminate" were combined into a binary variable alongside other shooting types, as exploratory data analysis in @fig-casualty-by-shooting-type revealed that these two categories are strongly associated with a higher number of casualties.
 
-### Alternative Models Considered
-- **Linear Regression**:
-  - Strengths: Simplicity and interpretability.
-  - Weaknesses: Inadequate for capturing non-linear relationships and interactions present in the data.
+Latitude is included as a predictor to account for geographic variability that might affect casualty likelihood.
+Longitude was evaluated during model selection but excluded due to its lack of statistical significance.
 
-- **Gradient Boosting Machines (GBM)**:
-  - Strengths: Often more accurate due to its iterative learning.
-  - Weaknesses: Higher risk of overfitting, more computationally expensive and less intepretibility.
+## Model Limitations {#sec-model-assumplim}
 
-The Random Forest model was chosen over these alternatives for its balance of flexibility, robustness, and interpretability of feature importance.
+The predictive performance of the Bayesian logistic regression model may be limited under certain circumstances.
+One key limitation is that the model's predictions might not generalize well if there are significant changes in societal or policy conditions, as these factors could alter the relationships between predictors and the outcome.
+Additionally, the model's reliability is sensitive to the quality of both the prior distributions and the dataset.
+Poorly chosen priors or biased and incomplete data can lead to inaccurate predictions and reduced robustness of the model.
 
-# Results
+## Model Validation {#sec-model-modelvalidation}
 
-Our results are summarized in @tbl-modelresults.
+The Bayesian logistic regression model is validated through several diagnostics.
+In @fig-ppcheckandposteriorvsprior-1, the posterior predictive check shows alignment between the posterior predictive distribution ($y_{rep}$) and the observed data ($y$), indicating a good fit.
+@fig-ppcheckandposteriorvsprior-2 shows that posterior estimates are narrower and shifted compared to priors, reflecting the influence of the data.
+Additionally, @fig-stanareyouokay-1 confirms well-mixed and stationary MCMC chains, while @fig-stanareyouokay-2 shows $\hat{R} \approx 1.00$ for all parameters, verifying convergence and reliability.
+
+## Alternative Models Considered {#sec-model-altmodel}
+
+Several models were evaluated during the development process:
+
+1. **Linear Logistic Regression**:
+   - **Strengths**: Simplicity and interpretability.
+   - **Weaknesses**: Cannot incorporate prior knowledge or provide uncertainty quantification.
+
+2. **Random Forest**:
+   - **Strengths**: Handles complex, non-linear interactions.
+   - **Weaknesses**: Limited interpretability and probabilistic outputs.
+
+The Bayesian logistic regression model was chosen for its ability to provide interpretable, probabilistic predictions while quantifying uncertainty and incorporating prior information.
+
+# Results {#sec-result}
+
+The model results are summarized in @tbl-modelresults.
+
+```{r}
+#| echo: false
+#| eval: true
+#| warning: false
+#| message: false
+#| label: tbl-modelresults
+#| tbl-cap: "Explanatory model of the Existence of Casualty in Each School Shootig Event based on Shooting Type and Latitude"
+
+# Load the saved models
+bayes_model<- readRDS(file = here::here("models/bayes_model.rds"))
+
+modelsummary(
+  list("Casualty Occurrence" = bayes_model),
+  fmt = 2,
+  statistic = c("({conf.low}, {conf.high})"),
+  gof_omit = "^(R2|RMSE|Log.Lik)",  # Omit metadata
+  output = "latex"
+)
+```
+
+## Intercept  
+The intercept estimate is **2.6798** (95% credible interval: **1.0774 to 4.3091**), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels.
+This corresponds to a high baseline probability, with the credible interval excluding zero, indicating strong evidence for a nonzero probability of casualties under these conditions.
+
+## Shooting Type  
+The coefficient for the "other" shooting category is **-0.6336** (95% credible interval: **-1.0735 to -0.1784**), suggesting a lower likelihood of casualties compared to the "indiscriminate" or "targeted" categories.
+
+## Latitude  
+The latitude coefficient is **-0.0443** (95% credible interval: **-0.0876 to -0.0021**), indicating a slight but significant decrease in the probability of casualties with increasing latitude.
+This suggests potential geographic trends influencing casualty outcomes, such as differences in population density, response capabilities, or regional policies.
+
+## Predicted Existence of Casualties by State {#sec-result-predicted}
+
+Using a Bayesian Logistic regression model, predictions for the existence of casualties in each state were generated.
+The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization.
 
@@
 #| label: fig-predicted_shooting_visualization
-#| tbl-cap: "Predicted 2025 US School Shootings Numbers"
+#| fig-cap: "Existence of Casualties in US School Shootings Prediction"
 #| warning: false
+#| width: 8
+#| height: 4
+# Load necessary libraries
+library(ggplot2)
+library(dplyr)
+library(maps)
+library(patchwork)
+library(rstanarm)
 
-#### Step 1: Load the Data and Model ####
-# Load cleaned data
-clean_data <- read.csv("../data/02-analysis_data/no_outlier_data.csv")
+us_states <- map_data("state")
 
-# Ensure 'state' is a factor
-clean_data$state <- as.factor(clean_data$state)
+# Calculate state centroids (mean latitude and longitude for each state)
+state_centroids <- us_states %>%
+  group_by(region) %>%
+  summarise(long = mean(range(long)), lat = mean(range(lat)))
 
-# Load the Random Forest model
-rf_model_clean <- readRDS(file = here::here("models/rf_model.rds"))
-
-#### Step 2: Prepare Data for Predictions ####
-# Year(s) for prediction
-years <- 2025
-
-# Create a data frame with unique states and years
-prediction_data <- expand.grid(
-  year = years,
-  state = unique(clean_data$state)  # Use unique states
+# Fix Florida's centroid directly
+state_centroids <- state_centroids %>%
+  mutate(
+    long = ifelse(region == "florida", -81.686783, long),
+    lat = ifelse(region == "florida", 27.766279, lat)
   )
 
-# Convert 'state' to a factor and match levels with training data
-prediction_data$state <- factor(prediction_data$state, levels = levels(clean_data$state))
-
-# Ensure no missing levels
-if (any(is.na(prediction_data$state))) {
-  stop("State levels in prediction_data do not match clean_data.")
-}
-
-# Make predictions
-predictions <- predict(rf_model_clean, newdata = prediction_data)
-
-# Add predictions to the data frame
-prediction_data$predicted_casualties <- predictions
-
-#### Step 3: Prepare Map Data ####
-# Get US map data
-us_map <- map_data("state")
-
-# Convert state names in `us_map` to lowercase
-us_map$region <- tolower(us_map$region)
-
-# Convert state in `prediction_data` to lowercase for compatibility
-prediction_data$state <- tolower(as.character(prediction_data$state))
-
-# Merge map data with predictions
-map_data <- us_map %>%
-  left_join(prediction_data, by = c("region" = "state"))
-
-# Fill missing predictions with 0 (optional, avoids grey states)
-map_data$predicted_casualties[is.na(map_data$predicted_casualties)] <- 0
-
-#### Step 4: Plot the Map ####
-ggplot(map_data, aes(long, lat, group = group, fill = predicted_casualties)) +
-  geom_polygon(color = "white") +
-  scale_fill_gradient(
-    low = "#E6F7FF",  # Light blue
-    high = "#0072B2", # Dark blue
-    name = "Casualties"
-  ) +
-  labs(
-    title = "Predicted US School Shooting Casualties by State (2025)",
-    subtitle = "Number of casualties visualized geographically",
-    x = NULL,
-    y = NULL
-  ) +
-  coord_fixed(1.3) +
-  theme_minimal(base_size = 14) +
-  theme(
-    plot.title = element_text(hjust = 0.5, face = "bold", size = 10),
-    plot.subtitle = element_text(hjust = 0.5, size = 10),
-    legend.position = "right",
-    axis.text = element_blank(),
-    axis.ticks = element_blank(),
-    panel.grid = element_blank()
+# Continue with the rest of your code to predict and plot as before
+state_data <- state_centroids %>%
+  mutate(
+    shooting_category = "indiscriminate or targeted",
+    lat = lat,
+    long = long
   )
 
+# Predict probabilities and classify shooting existence
+state_data$prob_targeted <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
+state_data$shooting_category <- "other"
+state_data$prob_other <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
+state_data$exist_targeted <- ifelse(state_data$prob_targeted > 0.70657, 1, 0)
+state_data$exist_other <- ifelse(state_data$prob_other > 0.70657, 1, 0)
+
+# Plot for "Targeted/Indiscriminate"
+p1 <- ggplot(us_states) +
+  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
+  geom_point(data = state_data %>% filter(exist_targeted == 1), 
+             aes(x = long, y = lat), 
+             color = "darkred", size = 4, alpha = 0.8) +
+  theme_void() +
+  labs(title = "Shooting Type: Targeted/Indiscriminate", x = NULL, y = NULL) +
+  theme(
+    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
+  )
+
+# Plot for "Other"
+p2 <- ggplot(us_states) +
+  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
+  geom_point(data = state_data %>% filter(exist_other == 1), 
+             aes(x = long, y = lat), 
+             color = "darkblue", size = 4, alpha = 0.8) +
+  theme_void() +
+  labs(title = "Shooting Type: Other", x = NULL, y = NULL) +
+  theme(
+    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
+  )
+
+# Combine plots with patchwork
+combined_plot <- p1 + p2 +
+  plot_annotation(
+    theme = theme(
+      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
+    )
+  )
+
+# Display the final plot
+print(combined_plot)
 ```
 
-```{r}
-#| echo: false
-#| eval: false
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
+# Discussion {#sec-discussion}
 
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+## Contributions of This Study  
+This paper utilizes a Bayesian Logistic Regression model to predict the likelihood of casualties in school shootings, focusing on the impact of shooting type and geographic latitude.
+The findings demonstrate that "Targeted" and "Indiscriminate" shootings significantly increase casualty probabilities, necessitating their classification into a unified predictive category.
+This categorization improves the model's predictive accuracy and highlights the distinct risk these shooting types pose.
+Furthermore, the model generates spatial predictions at the state level, identifying regions with elevated casualty risks.
+These predictions provide policymakers with concrete data to prioritize interventions and allocate resources effectively, especially in high-risk areas.
 
-# Discussion
+## Policy Implications
 
-## First discussion point {#sec-first-point}
+### Geographic Targeting of Resources  
+The geographic analysis in this study reveals significant disparities in casualty risks, with southern states showing higher probabilities of school shootings resulting in casualties.
+This finding emphasizes the need for tailored interventions in these regions.
+For example, policymakers can allocate funding to implement enhanced physical security measures in schools, such as metal detectors, reinforced entry points, and surveillance systems.
+Additionally, improving emergency response systems, including faster deployment of law enforcement and medical personnel, can mitigate the impact of shootings in high-risk areas.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+### Focus on High-Risk Shooting Types  
+The model's findings confirm that "Targeted" and "Indiscriminate" shootings are strongly associated with higher casualty rates.
+Policymakers can design prevention strategies specifically aimed at these shooting types.
+For "Targeted" shootings, schools can adopt proactive threat assessment programs to identify and provide support to at-risk individuals.
+For "Indiscriminate" shootings, community-level initiatives addressing factors such as mental health, social isolation, and access to firearms can reduce the likelihood of such incidents.
 
-## Second discussion point
+### Strengthening Gun Control Measures  
+Given the role of firearms in school shootings, this study highlights the importance of stricter gun control measures.
+Policies such as universal background checks, red flag laws, and safe firearm storage requirements can reduce unauthorized access to firearms and lower casualty risks.
+These measures can be particularly impactful in regions identified as having elevated risks, complementing other targeted interventions.
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+### Investment in Mental Health Resources  
+The findings also point to the importance of addressing underlying causes, such as untreated mental health conditions, that may contribute to school shootings.
+Policymakers can expand access to mental health services by increasing funding for school counselors and psychologists, creating community-based support programs, and implementing mental health education initiatives.
+These efforts can reduce risk factors while fostering a supportive environment for students.
 
-## Third discussion point
+### Ongoing Policy Evaluation  
+Effective policymaking requires regular assessment of interventions.
+This study provides a framework for evaluating the impact of existing measures, such as increased school security or stricter firearm regulations.
+Policymakers can use predictive models to monitor changes in casualty risks and adapt strategies based on emerging data.
+For example, periodic updates to the model with new data can refine predictions and improve policy responsiveness.
 
-## Weaknesses and next steps
+## Limitations of the Analysis  
+The reliability of the model depends on the availability and quality of the underlying data and priors.
+Changes in societal conditions or policy environments may reduce the model's predictive power over time.
+Additionally, the analysis does not account for several important contextual factors, such as individual school characteristics, socio-economic disparities, and differences in firearm accessibility across states.
+Including these factors could improve the model's ability to capture the complexity of school shooting dynamics.
 
-Weaknesses and next steps should also be included.
+## Recommendations for Future Research
+
+### Expanding the Predictive Framework  
+Future studies should incorporate additional predictors to enhance the model’s scope and accuracy.
+Factors such as school size, local economic conditions, and access to firearms could provide a more detailed understanding of the drivers of casualties.
+Incorporating longitudinal data could also help capture temporal trends and the impact of policy changes over time.
+
+### Evaluating Policy Effectiveness  
+To ensure that interventions are impactful, future research should evaluate the effectiveness of existing policies.
+For example, studies could assess whether stricter gun control laws or increased mental health funding significantly reduce the likelihood or severity of school shootings.
+Policymakers could use these insights to refine strategies and develop evidence-based policies.
+
+### Encouraging Interdisciplinary Collaboration  
+Addressing the complex issue of school shootings requires collaboration across disciplines.
+Future research should bring together experts from education, mental health, public policy, and law enforcement to design holistic interventions.
+Collaborative efforts can also facilitate data sharing, improve predictive modeling, and create a more comprehensive understanding of how to reduce casualties effectively.
 
@@
 
-# Appendix {-}
+# Appendix {#sec-appendix}
 
-# Additional data details
+## Additional data details
 
-# Model details {#sec-model-details}
+## Model details {#sec-model-details}
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
+In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check.
+This shows that the posterior predictive distribution ($y_{rep}$) aligns well with the observed data ($y$), indicating that the model provides a reasonable fit to the data and captures its underlying structure effectively.
 
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
+In @fig-ppcheckandposteriorvsprior-2, we compare the posterior with the prior distributions.
+This shows that the posterior estimates for all parameters ($Intercept$, $lat$, and $shooting\_categoryother$ are informed by the data, as the posterior distributions are narrower and shifted relative to the priors, reflecting updated beliefs based on the observed evidence.
 
 ```{r}
-#| eval: false
+#| eval: true
 #| echo: false
@@
 
-pp_check(first_model) +
+pp_check(bayes_model) +
   theme_classic() +
@@
 
-posterior_vs_prior(first_model) +
+posterior_vs_prior(bayes_model) +
   theme_minimal() +
@@
 
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
+@fig-stanareyouokay-1 is a trace plot.
+It shows the sampling behavior of the MCMC algorithm for the model parameters (Intercept, shooting_categoryother, and lat) across four chains.
+The chains are well-mixed, stationary, and overlap significantly, indicating good convergence and effective exploration of the posterior distributions.
+It suggests that the model's parameters were sampled effectively, with no immediate signs of convergence issues.
 
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
+@fig-stanareyouokay-2 is a Rhat plot.
+This $\hat{R}$ plot shows that all parameters have $\hat{R} \approx 1.00$, indicating excellent convergence of the MCMC chains.
+The results suggest that the chains are well-mixed, and the posterior distributions can be considered reliable for interpretation.
 
@@
 #| echo: false
-#| eval: false
+#| eval: true
 #| message: false
@@
 
-plot(first_model, "trace")
+plot(bayes_model, "trace")
 
-plot(first_model, "rhat")
+plot(bayes_model, "rhat")
 ```

### 2024-12-03 scripts/03-clean_data.R

@@ -131,4 +131,3 @@ head(school_shoot_cleaned)
 
 #### Save data ####
 write_parquet(school_shoot_cleaned, "data/02-analysis_data/analysis_data.parquet")
-write_csv(school_shoot_cleaned, "data/02-analysis_data/analysis_data.csv")

### 2024-12-03 scripts/04-test_analysis_data.R

@@ -14,7 +14,7 @@ library(dplyr)
 library(testthat)
 library(arrow)
 
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+cleaned_data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 
 #### Test data ####
@@ -56,19 +56,8 @@ test_cleaned_data <- function(data) {
     }
   }
   
-  # 4. Validate `casualties` consistency
-  cat("\n--- Casualties Consistency ---\n")
-  casualties_discrepancies <- data %>%
-    filter(casualties != (killed + injured)) %>%
-    select(uid, killed, injured, casualties)
-  if (nrow(casualties_discrepancies) > 0) {
-    cat("ERROR: Discrepancies in `casualties` column:\n")
-    print(casualties_discrepancies)
-  } else {
-    cat("`casualties` column is consistent.\n")
-  }
   
-  # 5. Check date column integrity
+  # 4. Check date column integrity
   cat("\n--- Date Validation ---\n")
   if ("date" %in% names(data)) {
     invalid_dates <- data %>%
@@ -81,7 +70,7 @@ test_cleaned_data <- function(data) {
     }
   }
   
-  # 6. Check for duplicates
+  # 5. Check for duplicates
   cat("\n--- Duplicate Check ---\n")
   duplicate_rows <- data %>%
     duplicated()
@@ -92,13 +81,13 @@ test_cleaned_data <- function(data) {
     cat("No duplicate rows found.\n")
   }
   
-  # 7. Summary statistics for numeric columns
+  # 6. Summary statistics for numeric columns
   cat("\n--- Summary Statistics ---\n")
   numeric_columns <- data %>%
     select(where(is.numeric))
   print(summary(numeric_columns))
   
-  # 8. Validate categorical columns
+  # 7. Validate categorical columns
   cat("\n--- Categorical Column Validation ---\n")
   categorical_columns <- c("school_type", "state", "shooting_type", "gender_shooter1")
   for (col in categorical_columns) {
@@ -112,4 +101,5 @@ test_cleaned_data <- function(data) {
 }
 
 # Test the cleaned dataset
-test_cleaned_data(school_shoot_cleaned)
+test_cleaned_data(cleaned_data)
+

### 2024-12-03 scripts/05-exploratory_data_analysis.R

@@ -10,8 +10,10 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(arrow)
+
 #### Read data ####
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
+data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
 # Dimensions of the dataset
 dim(data)
@@ -50,7 +52,6 @@ ggplot(state_counts, aes(x = reorder(state, count), y = count)) +
   coord_flip() +
   labs(title = "Number of Incidents by State", x = "State", y = "Count") +
   theme_minimal()
-
 
 # Causaulity Distribution
 ggplot(data, aes(x = casualties)) +

### 2024-12-03 scripts/06-model_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
+# Purpose: Models the relationship between the existence of casualty and shooting type and longtitude
 # Author: Yun Chu
-# Date: 25 November 2024
+# Date: 30 November 2024
 # Contact: yun.chu@mail.utoronto.ca
 # License: CC BY-NC-SA 4.0
 # Pre-requisites: 03-clean_data.R has been run
@@ -11,112 +11,101 @@
 #### Workspace setup ####
 library(tidyverse)
 library(dplyr)
-library(randomForest)
-
-#### Read data ####
-data <- read_csv("data/02-analysis_data/analysis_data.csv")
-
+library(rstanarm)
+library(pROC)
+library(arrow)
+library(bayesplot)
 ### Model data ####
 
-# Add year column and aggregate data by state and year
-data <- data %>%
-  group_by(state, year) %>%
-  summarise(
-    killed = sum(killed, na.rm = TRUE),
-    injured = sum(injured, na.rm = TRUE),
-    casualties = sum(casualties, na.rm = TRUE)
-  ) %>%
-  ungroup()
+data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
-# Prepare data for modeling
-train_data <- data %>% filter(year <= 2024)
+# Step 1: Create a binary response variable
+data$ifcasualty <- ifelse(data$casualties> 0, 1, 0)
 
-# Convert state to a factor (categorical variable)
-train_data <- train_data %>%
-  mutate(state = as.factor(state))
+# Create a new column with simplified categories
+data$shooting_category <- ifelse(
+  data$shooting_type %in% c("indiscriminate", "targeted"),  
+  "indiscriminate or targeted",                              
+  "other"
+)
 
-# Define features and target variable
-X <- train_data %>% select(year, state)
-y <- train_data$casualties
+data <- na.omit(data[, c("ifcasualty", "shooting_category", "long", "lat")])
 
-#### Cross-Validation for Residual Calculation ####
+#posterior <- as.matrix(bayes_model)
+#hist(posterior[, "shooting_categoryother"], main = "Posterior Distribution")
 
-# Define cross-validation folds
-library(caret)
-set.seed(42)
-folds <- createFolds(y, k = 5, list = TRUE)
+# Fit Bayesian logistic regression
[…]

### 2024-12-03 scripts/07-replications.R

@@ -1,19 +1,0 @@
- #### Preamble ####
-# Purpose: Replicated graphs from... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
-# License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(tidyverse)
-# [...UPDATE THIS...]
-
-#### Load data ####
-# [...ADD CODE HERE...]
-
-
-

### 2024-12-03 paper/paper.qmd
add datasheet & Surveys, Sampling and Observational Data sections

@@
 The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] .
-The following libraries are utilized in this paper:
-
-- tidyverse [@tidyverse2024]
-- dplyr [@dplyr2024]
-- lubridate [@lubridate2024]
-- readr [@readr2024]
-- stringr [@stringr2024]
-- arrow [@arrow2024]
-- testthat [@testthat2024]
-- modelsummary [@modelsummary]
-- ggplot2 [@ggplot2]
-- maps [@maps]
-- knitr [@knitr]
-- here [@here]
-- kableExtra [@kableExtra]
-- rstanarm [@rstanarm]
+The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm].
 
@@
 
+The Washington Post spent a year determining how many children have been affected by school shootings, beyond just those killed or injured.
 The dataset, compiled by The Washington Post, translates real-world school shooting incidents into structured entries by aggregating information from news articles, open-source databases, law enforcement reports, and direct calls to schools.
@@
 
-## Additional data details
+## **Analysis of The Washington Post's Data Collection Approach**
+
+The Washington Post’s investigation into school shootings exemplifies a meticulous approach to data collection and analysis, aligning with best practices in data journalism and social science research.
+
+### **Data Collection and Triangulation**
+The Post utilized multiple data sources—including Nexis, news articles, open-source databases, law enforcement reports, and direct communications with schools and police departments—to compile a comprehensive list of school shootings.
+This method reflects the principle of data triangulation, which enhances the validity of findings by cross-verifying information from diverse sources [@denzin1978research].
+Such triangulation is crucial in data journalism to ensure accuracy and comprehensiveness.
+
+### **Defining Inclusion and Exclusion Criteria**
+By establishing clear criteria for what constitutes a school shooting—excluding incidents like after-hours events, accidental discharges without injury, and private suicides—the Post ensured a focused analysis.
+This practice aligns with the concept of construct validity, ensuring that the study accurately reflects the phenomenon under investigation [@cronbach1955construct].
+
+### **Sampling and Estimation Techniques**
+To estimate the number of children exposed to gunfire, the Post used enrollment figures from the U.S.
+Education Department and adjusted for average daily absenteeism.
+Specifically, they deducted 7% from enrollment totals to account for average absenteeism and reduced enrollment by 50% for shootings occurring just before or after school hours.
+This approach mirrors sampling techniques discussed in "Telling Stories with Data", which emphasizes the importance of identifying the target population, understanding the context-specific variables (such as absenteeism rates), and applying appropriate adjustments to ensure the representativeness and reliability of the results [@alexander2023telling].
+For example, the text highlights the necessity of incorporating real-world constraints—such as data availability or measurement gaps—into the design of sampling frameworks to improve the accuracy of population estimates.
+
+### **Reproducibility and Transparency**
+The Post's transparency in detailing their methodology allows for reproducibility and critical assessment, key components in data journalism.
+As highlighted in *Telling Stories with Data*, reproducible workflows involve sharing data, code, and analytical steps, enabling others to replicate and validate findings [@alexander2023telling].
+
+### **Addressing Potential Biases**
+The Post acknowledged that their database might not include every qualifying incident due to the absence of a federal tracking system.
+This recognition of potential data availability bias is essential, as discussed in the literature on observational data, which emphasizes the importance of considering who is included in the dataset and who is systematically excluded [@kaplan2014bias].
+
+#### **Weaknesses**
+Despite its strengths, The Washington Post’s methodology has notable limitations.
+Reliance on media reports and open-source databases may introduce reporting bias, potentially underrepresenting incidents in areas with limited media coverage [@kaplan2014bias].
+The exclusion of certain incidents, such as private suicides, could understate the broader psychological impact of gun violence.
+Additionally, adjustments for absenteeism rates assume consistent patterns across all schools, which may overlook localized variations and specific community contexts [@alexander2023telling].
+The use of proprietary tools like Nexis and reliance on non-public datasets also pose challenges for reproducibility, limiting the ability of others to independently verify findings.
+Finally, the absence of a centralized federal tracking system for school shootings creates a systemic data gap that even the Post’s robust methodology cannot fully address [@kaplan2014bias].
+
+### **Conclusion**
+The Washington Post's methodology in investigating school shootings demonstrates a rigorous application of data collection and analysis principles, aligning with best practices in data journalism and social science research.
+By employing data triangulation [@denzin1978research], defining clear inclusion criteria [@cronbach1955construct], adjusting for sampling considerations [@alexander2023telling], ensuring reproducibility [@alexander2023telling], and acknowledging potential biases [@kaplan2014bias], the Post provides a robust framework for understanding the impact of school shootings on children.
+Nonetheless, the identified weaknesses underscore the importance of continued refinement and broader systemic solutions for data collection.
 

### 2024-12-03 paper/paper.qmd
update pictures and scripts

@@
 ---
-title: "2025 US School Shooting Predictions and Solutions"
-subtitle: "Leveraging Forecasting to Shape Policy and Protect Communities"
+title: "Understanding the Risk of Casualties in US School Shootings"
+subtitle: "Leveraging Bayesian Analysis to Guide Safety and Prevention Strategies"
 author: 
@@
 date-format: long
-abstract: "School shootings in the United States pose a critical challenge with far-reaching impacts.
-This paper employs a Random Forest regression model to predict state-level casualties for 2025 using historical data from 1999 to the present.
-The findings reveal a positively skewed distribution of casualties, with states like Georgia, California, and Pennsylvania projected to face the highest numbers.
-These insights highlight the need for targeted interventions, including enhanced school safety measures and stricter gun control policies.
-By providing actionable predictions, this study aims to inform policies that reduce school shooting casualties and protect vulnerable populations."
+abstract: "School shootings in the United States are a persistent and devastating issue.
+This paper utilizes a Bayesian Logistic Regression model to predict the probability of casualties in school shooting incidents, using historical data from 1999 onward.
+Key predictors, including shooting type and geographic latitude, are identified and analyzed.
+The results demonstrate that targeted and indiscriminate shootings significantly increase casualty risks, highlighting the need for targeted safety measures, such as enhanced threat assessment programs and stricter firearm regulations.
+This analysis provides a framework for implementing data-driven strategies to reduce casualties and improve school safety nationwide."
 format: pdf
@@
 library(arrow)
+library(patchwork)
+library(rstanarm)
 #### Read data ####
 data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
-data$ifcasualty <- ifelse(data$casualties> 0, 1, 0)
-# Create a new column with simplified categories
-data$shooting_category <- ifelse(
-  data$shooting_type %in% c("indiscriminate", "targeted"),  
-  "indiscriminate or targeted",                              
-  "other"
-)
-
 ```
@@
 The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] .
-The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm].
+The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm], `arrow` [@R-arrow] and `patchwork` [@patchwork].
 
@@
 #| label: fig-casualty-by-shooting-type
-#| fig-cap: "Distribution of Casualty By Shooting Type"
+#| fig-cap: "Number of Casualties for Each Shooting Type in US School Shooting Incidents"
 #| echo: false
@@
   labs(
-    title = "Number of Casualties by Shooting Type",
+    title = "Distribution of Casualty By Shooting Type",
     x = "Shooting Type",
@@
 #| echo: false
-#| fig-cap: Distribution of Casualties
+#| fig-cap: "Distribution of Casualties in US School Shooting Incidents from 1999 to 2024"
 # Causaulity Distribution
@@
   geom_histogram(binwidth = 1, fill = "skyblue", color = "white") + 
-  labs(x = "Number of Casualties", y = "Frequency") +
-  theme_minimal()
+  labs(title = "Distribution of Casualties", x = "Number of Casualties", y = "Frequency") +
+  theme_minimal() +
+    theme(
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
+    )
 
@@
 #| label: fig-Shooting_Casaulties_by_State
-#| fig-cap: "US School Shooting Casaulties by State"
+#| fig-cap: "US School Shooting Casaulties by State from 1999 to 2024"
 #| echo: false
@@
 
-# Plot the map
+# Adjusted Plot
 ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
@@
   scale_fill_gradient(
-    low = "#E6F7FF",  # Light blue
-    high = "#0072B2",  # Dark blue
+    low = "#E6F7FF",
+    high = "#0072B2",
     name = "Total Casualties",
@@
   labs(
-    title = "US School Shooting Casualties by State",
-    subtitle = "Number of casualties visualized geographically",
+    title = "US School Shooting Casualties Visualization",
     x = NULL,
@@
   theme(
-    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 13, margin = ggplot2::margin(t = 10, b = 10)),
-    plot.subtitle = ggplot2::element_text(hjust = 0.5, size = 10, margin = ggplot2::margin(t = 0, b = 20)),
-    legend.position = "right",  # Alternative: Use "bottom" for a horizontal legend
-    legend.key.height = unit(1, "cm"),  # Adjust legend size
-    axis.text = ggplot2::element_blank(),
-    axis.ticks = ggplot2::element_blank(),
-    panel.grid = ggplot2::element_blank(),
-    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Increase margins for PDF export
+    plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(t = 10, b = 10)),
+    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(t = 0, b = 20)),
+    legend.position = "right",
+    legend.title = element_text(size = 10),  # Reduce the size of "Total Casualties"
+    legend.text = element_text(size = 8),   # Reduce size of legend values
+    legend.key.height = unit(0.5, "cm"),    # Reduce the height of the legend color bar
+    legend.key.width = unit(0.25, "cm"),    # Reduce the width of the legend color bar
+    axis.text = element_blank(),
+    axis.ticks = element_blank(),
+    panel.grid = element_blank(),
+    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)  # Increase margins for spacing
+  ) +
+  guides(
+    fill = guide_colorbar(
+      barwidth = 0.5,  # Adjust the width of the legend bar
+      barheight = 8    # Adjust the height of the legend bar
+    )
   )
@@
 #| label: fig-shootingtypedist
-#| fig-cap: "Shooting Type Distribution" 
+#| fig-cap: "US School Shooting Type Distribution from 1999 to 2024" 
 #| echo: false
 #| eval: true
-
-# Load necessary libraries
-library(ggplot2)
-library(dplyr)
-library(ggrepel)
 
@@
   mutate(
-    proportion = n / sum(n),  # Calculate proportions
-    label = paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")")  # Labels
+    proportion = n / sum(n)  # Calculate proportions
   )
 
-# Calculate label positions and angles
-shooting_data <- shooting_data %>%
-  mutate(
-    y_position = cumsum(proportion) - 0.5 * proportion,  # Calculate mid-point of each segment
-    angle = 90 - (360 * (cumsum(proportion) - proportion / 2))  # Adjust angles for better label alignment
-  )
-
-# Plot Pie Chart
+# Plot Pie Chart without Text
 ggplot(shooting_data, aes(x = "", y = proportion, fill = shooting_type)) +
@@
   ) +
-  geom_text(
-    aes(
-      label = ifelse(proportion > 0.05, paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")"), ""),
-      y = y_position,
-      angle = ifelse(proportion < 0.1, angle, 0)  # Add dynamic rotation for small segments
-    ),
-    size = 3,
-    hjust = ifelse(shooting_data$proportion < 0.1, 1, 0.5),  # Align labels for smaller slices
-    nudge_x = 0.4  # Nudge outward for better alignment
-  ) +
   theme(
@@
   )
+
 ```
@@
 #| label: fig-latitudedist
-#| fig-cap: "Distribution of Latitude of Schools in School Shooting Events"
+#| fig-cap: "Distribution of Latitude of Schools in School Shooting Incidents in US from 1999 to 2024"
 #| echo: false
@@
 #| warning: false
-
-# Load necessary libraries
-library(ggplot2)
-
-# Assuming `data` contains the `lat` column (latitude values)
-# Replace `data` with your actual dataset
 
@@
   labs(
-    title = "Distribution of Latitude",
+    title = "Distribution of Latitude of Schools in Shooting Incidents",
     x = "Latitude",
@@
 ## Intercept  
-The intercept estimate is **2.6798** (95% credible interval: **1.0774 to 4.3091**), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels.
+The intercept estimate is 2.37 (95% credible interval: 0.824 to 4.02), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels.
 This corresponds to a high baseline probability, with the credible interval excluding zero, indicating strong evidence for a nonzero probability of casualties under these conditions.
@@
 ## Shooting Type  
-The coefficient for the "other" shooting category is **-0.6336** (95% credible interval: **-1.0735 to -0.1784**), suggesting a lower likelihood of casualties compared to the "indiscriminate" or "targeted" categories.
+The coefficient for the "other" shooting category is -0.63 (95% credible interval: -1.09 to -0.16), suggesting a lower likelihood of casualties compared to the "indiscriminate" or "targeted" categories.
 
 ## Latitude  
-The latitude coefficient is **-0.0443** (95% credible interval: **-0.0876 to -0.0021**), indicating a slight but significant decrease in the probability of casualties with increasing latitude.
+The latitude coefficient is -0.04 (95% credible interval: -0.08 to 0.00), indicating a slight but significant decrease in the probability of casualties with increasing latitude.
 This suggests potential geographic trends influencing casualty outcomes, such as differences in population density, response capabilities, or regional policies.
@@
 Using a Bayesian Logistic regression model, predictions for the existence of casualties in each state were generated.
-The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization.
+The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization where the states in dark red are classified as casualty exists while the states in grey are classified as no casualty.
 
@@
 #| label: fig-predicted_shooting_visualization
-#| fig-cap: "Existence of Casualties in US School Shootings Prediction"
+#| fig-cap: "Prediction of the Existence of Casualty in US School Shootings Incidents Based on Bayesian Logistic Regression Model"
 #| warning: false
-#| width: 8
-#| height: 4
-# Load necessary libraries
-library(ggplot2)
-library(dplyr)
-library(maps)
-library(patchwork)
-library(rstanarm)
 
+# Load US map data
 us_states <- map_data("state")
 
-# Calculate state centroids (mean latitude and longitude for each state)
-state_centroids <- us_states %>%
-  group_by(region) %>%
-  summarise(long = mean(range(long)), lat = mean(range(lat)))
-
-# Fix Florida's centroid directly
-state_centroids <- state_centroids %>%
+# Prepare state-level data for predictions
+state_data <- us_states %>%
+  distinct(region) %>%  # Get unique state names
   mutate(
-    long = ifelse(region == "florida", -81.686783, long),
-    lat = ifelse(region == "florida", 27.766279, lat)
+    shooting_category = "indiscriminate or targeted" # Initial shooting category for predictions
   )
 
-# Continue with the rest of your code to predict and plot as before
-state_data <- state_centroids %>%
-  mutate(
-    shooting_category = "indiscriminate or targeted",
-    lat = lat,
-    long = long
+# Add latitude for each state by calculating the mean latitude of the state polygons
+state_data <- state_data %>%
+  left_join(
+    us_states %>%
+      group_by(region) %>%
+      summarise(lat = mean(lat)),  # Calculate mean latitude for each state
+    by = "region"
   )
 
-# Predict probabilities and classify shooting existence
+# Predict probabilities using the Bayesian model
 state_data$prob_targeted <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
-state_data$shooting_category <- "other"
+state_data$shooting_category <- "other"  # Change category for next prediction
 state_data$prob_other <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
-state_data$exist_targeted <- ifelse(state_data$prob_targeted > 0.70657, 1, 0)
-state_data$exist_other <- ifelse(state_data$prob_other > 0.70657, 1, 0)
+
+# Classify shooting existence based on thresholds
+state_data <- state_data %>%
+  mutate(
+    exist_targeted = ifelse(prob_targeted > 0.70657, 1, 0),
+    exist_other = ifelse(prob_other > 0.70657, 1, 0)
+  )
+
+# Join predictions with the US map data
+map_data <- us_states %>%
+  left_join(state_data, by = "region") %>%  # Merge predictions with state polygons
+  rename(lat = lat.x)  # Rename lat.x back to lat for compatibility with ggplot
 
 # Plot for "Targeted/Indiscriminate"
-p1 <- ggplot(us_states) +
-  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
-  geom_point(data = state_data %>% filter(exist_targeted == 1), 
-             aes(x = long, y = lat), 
-             color = "darkred", size = 4, alpha = 0.8) +
+p1 <- ggplot(map_data) +
+  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(exist_targeted)), color = "white") +
+  scale_fill_manual(values = c("0" = "lightgray", "1" = "darkred"), na.translate = FALSE) +
+  coord_fixed(1.3) +  # Fix aspect ratio
   theme_void() +
-  labs(title = "Shooting Type: Targeted/Indiscriminate", x = NULL, y = NULL) +
+  labs(
+    title = "Shooting Type: Targeted/Indiscriminate",
+    fill = NULL
+  ) +
   theme(
-    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
+    plot.title = element_text(size = 9, hjust = 0.5),
+    plot.margin = margin(t = 5, r = 10, b = 5, l = 10)  # Add margins for spacing
   )
@@
 # Plot for "Other"
-p2 <- ggplot(us_states) +
-  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
-  geom_point(data = state_data %>% filter(exist_other == 1), 
-             aes(x = long, y = lat), 
-             color = "darkblue", size = 4, alpha = 0.8) +
+p2 <- ggplot(map_data) +
+  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(exist_other)), color = "white") +
+  scale_fill_manual(values = c("0" = "lightgray", "1" = "darkblue"), na.translate = FALSE) +
+  coord_fixed(1.3) +  # Fix aspect ratio
   theme_void() +
-  labs(title = "Shooting Type: Other", x = NULL, y = NULL) +
+  labs(
+    title = "Shooting Type: Other",
+    fill = NULL
+  ) +
   theme(
-    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
+    plot.title = element_text(size = 9, hjust = 0.5),
+    plot.margin = margin(t = 5, r = 10, b = 5, l = 10)  # Add margins for spacing
   )
 
-# Combine plots with patchwork
+# Combine the two plots using patchwork and add a title
 combined_plot <- p1 + p2 +
   plot_annotation(
+    title = "Predicted Existence of Casualties Visualization",
     theme = theme(
-      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
+      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
+      plot.margin = margin(t = 5, r = 5, b = 5, l = 5)  # Reduce margins around the combined plot
     )
@@
 
-# Display the final plot
+# Display the combined plot
 print(combined_plot)
+
 ```
@@
 
-## **Analysis of The Washington Post's Data Collection Approach**
+## Analysis of The Washington Post's Data Collection Approach
 
@@
 
-### **Data Collection and Triangulation**
+### Data Collection and Triangulation
 The Post utilized multiple data sources—including Nexis, news articles, open-source databases, law enforcement reports, and direct communications with schools and police departments—to compile a comprehensive list of school shootings.
@@
 
-### **Defining Inclusion and Exclusion Criteria**
+### Defining Inclusion and Exclusion Criteria
 By establishing clear criteria for what constitutes a school shooting—excluding incidents like after-hours events, accidental discharges without injury, and private suicides—the Post ensured a focused analysis.
@@
 
-### **Sampling and Estimation Techniques**
+### Sampling and Estimation Techniques
 To estimate the number of children exposed to gunfire, the Post used enrollment figures from the U.S.
@@
 
-### **Reproducibility and Transparency**
+### Reproducibility and Transparency
 The Post's transparency in detailing their methodology allows for reproducibility and critical assessment, key components in data journalism.
@@
 
-### **Addressing Potential Biases**
+### Addressing Potential Biases
 The Post acknowledged that their database might not include every qualifying incident due to the absence of a federal tracking system.
@@
 
-#### **Weaknesses**
+### Weaknesses
 Despite its strengths, The Washington Post’s methodology has notable limitations.
@@
 
-### **Conclusion**
+### Conclusion
 The Washington Post's methodology in investigating school shootings demonstrates a rigorous application of data collection and analysis principles, aligning with best practices in data journalism and social science research.

### 2024-12-03 scripts/03-clean_data.R

@@ -11,123 +11,24 @@
 
 #### Workspace setup ####
 library(dplyr)
-library(lubridate)
-library(stringr)
 library(arrow)
 
 #### Clean data ####
 
-# Load the dataset
-school_shoot_raw_data <- read.csv("data/01-raw_data/raw_data.csv")
+raw_data <- read.csv("data/01-raw_data/raw_data.csv")
 
-# Column explanations from the provided file
-column_explanations <- list(
-  uid = "Unique identifier",
-  nces_school_id = "National Center for Education Statistics unique identifier for the school",
-  school_name = "Name of the school",
-  nces_district_id = "National Center for Education Statistics unique identifier for the school district",
-  district_name = "Name of the school district",
-  city = "City where the school is located",
-  state = "State where the school is located",
-  enrollment = "Total enrollment of the school",
-  killed = "Number of individuals killed during the incident",
-  injured = "Number of individuals injured during the incident",
-  date = "Date of the incident"
+# Work on a copy of the raw data to ensure raw data integrity
+analysis_data <- raw_data
+
+# Add a column indicating whether there were casualties
+analysis_data$ifcasualty <- ifelse(analysis_data$casualties > 0, 1, 0)
+
+# Create a new column with simplified shooting categories
+analysis_data$shooting_category <- ifelse(
+  analysis_data$shooting_type %in% c("indiscriminate", "targeted"),
+  "indiscriminate or targeted",
+  "other"
 )
 
-# Step 1: Inspect the dataset
-str(school_shoot_raw_data)
-summary(school_shoot_raw_data)
-head(school_shoot_raw_data)
-
-# Step 2: Rename columns for consistency
-school_shoot_cleaned <- school_shoot_raw_data %>%
-  rename(
-    uid = uid,
-    nces_school_id = nces_school_id,
-    school_name = school_name,
-    nces_district_id = nces_district_id,
-    district_name = district_name,
-    city = city,
-    state = state,
-    enrollment = enrollment,
-    killed = killed,
-    injured = injured,
-    date = date
-  )
-
-# Parse and clean the date column
-school_shoot_cleaned <- school_shoot_raw_data %>%
-  mutate(
-    # Parse date from "Month Day, Year" format
-    date = mdy(date),
-    # Extract year, month, and day
-    year = year(date),
[…]

### 2024-12-03 scripts/06-model_data.R

@@ -12,27 +12,13 @@
 library(tidyverse)
 library(dplyr)
 library(rstanarm)
-library(pROC)
 library(arrow)
-library(bayesplot)
 ### Model data ####
 
 data <- read_parquet("data/02-analysis_data/analysis_data.parquet")
 
-# Step 1: Create a binary response variable
-data$ifcasualty <- ifelse(data$casualties> 0, 1, 0)
-
-# Create a new column with simplified categories
-data$shooting_category <- ifelse(
-  data$shooting_type %in% c("indiscriminate", "targeted"),  
-  "indiscriminate or targeted",                              
-  "other"
-)
-
 data <- na.omit(data[, c("ifcasualty", "shooting_category", "long", "lat")])
 
-#posterior <- as.matrix(bayes_model)
-#hist(posterior[, "shooting_categoryother"], main = "Posterior Distribution")
 
 # Fit Bayesian logistic regression
 bayes_model <- stan_glm(ifcasualty ~ shooting_category + lat,
@@ -42,70 +28,10 @@ bayes_model <- stan_glm(ifcasualty ~ shooting_category + lat,
                         prior_intercept = normal(0, 5),
                         chains = 4, iter = 2000)
 
-# Summary of the model
-summary(bayes_model)
-
-# Extract posterior probabilities
-predicted_probs <- posterior_epred(bayes_model)
-data$predicted_prob_bayes <- colMeans(predicted_probs)
-
-# Generate the ROC object
-roc_obj <- roc(data$ifcasualty, data$predicted_prob_bayes)
-
-
-# Calculate the optimal threshold using Youden's Index
-optimal_coords <- coords(roc_obj, "best", ret = c("threshold", "specificity", "sensitivity"))
-
-
-# Extract the optimal threshold value from coords()
-optimal_threshold <- as.numeric(optimal_coords["threshold"])
-
-# Display the optimal threshold
-cat("Optimal Threshold:", optimal_threshold, "\n")
-
-# Classify using the optimal threshold
-data$predicted_class_bayes <- ifelse(data$predicted_prob_bayes > optimal_threshold, 1, 0)
-
-# Confusion matrix
-conf_matrix_bayes <- table(Predicted = data$predicted_class_bayes, Actual = data$ifcasualty)
-print(conf_matrix_bayes)
-
-# Accuracy
-accuracy_bayes <- sum(diag(conf_matrix_bayes)) / sum(conf_matrix_bayes)
[…]

### 2024-12-04 paper/paper.qmd
prrofread error correction

@@
 
-School shootings remain a grave concern in the United States, with devastating consequences for students, educators, and communities.
-Despite the growing public discourse and policy debates surrounding school safety, little progress has been made in effectively predicting and preventing these tragic incidents (Jonson, 2017; RAND Corporation, 2015).
-Understanding the patterns and predictors of school shootings is essential for designing evidence-based interventions to protect students and reduce casualties.
-This paper contributes to this critical issue by leveraging data-driven methodologies to analyze and predict the existence of school shooting casualties, with the aim of informing policy and prevention efforts.
+School shootings in the United States remain a deeply concerning issue, resulting in devastating consequences for students, educators, and communities.
+Despite significant public discourse and legislative efforts, effective strategies to predict and prevent such incidents remain underdeveloped [@rand2015school].
+Understanding the factors contributing to school shooting casualties is critical for designing evidence-based policies that address the root causes of this violence.
 
-The estimand in this paper is to classify whether casualty exists in a school shooting incident in US.
-Using historical data from The Washington Post [@washingtonpost_school_shootings], which provides a comprehensive record of school shootings, a Bayesian Logistic Regression model was developed to predict the existence of casualty based on shooting type and latitude of the school.
-The analysis includes summary statistics, spatial visualizations, and predictive modeling, providing actionable insights for policymakers.
-This study fills a critical gap in understanding the probability of school shooting casualties and offers data-driven recommendations for targeted interventions.
+The estimand in this paper is the casualty in a school shooting incident in US.
+Using data from The Washington Post [@washingtonpost_school_shootings], this paper seeks to address this gap by employing a Bayesian Logistic Regression model to analyze the probability of casualties in school shooting incidents.
+Using historical data from 1999 onward, the analysis focuses on key predictors, such as shooting type and geographic latitude, to uncover patterns influencing casualty risks.
+Unlike previous research, this study emphasizes predictive modeling to offer actionable recommendations for policy and intervention strategies.
 
-A key finding of this study is the uneven distribution of casualties across states, with some states, such as Georgia, California, and Pennsylvania, predicted to experience significantly higher casualties in 2025.
-This highlights the importance of state-specific policies and resource allocation.
-The analysis also reveals that the distribution of casualties is positively skewed, with most incidents involving relatively few casualties, but a small number of high-casualty events disproportionately contributing to the overall impact.
-This information underscore the need for targeted strategies to mitigate the effects of high-casualty incidents.
+The findings indicate that targeted and indiscriminate shootings notably raise the probability of casualties, while geographic latitude offers further understanding of regional patterns.
+This analysis underscores the importance of tailored interventions, including proactive threat assessment programs and stricter firearm regulations, to mitigate the risks associated with high-casualty incidents.
 
-The importance of this study lies in its ability to provide a data-driven foundation for policy and prevention efforts.
-By identifying high-risk states and understanding the characteristics of high-casualty incidents, this paper aims to guide policymakers in allocating resources effectively and designing interventions that address the underlying factors contributing to school shootings.
+This study is significant because it provides policymakers with a data-driven framework to allocate resources effectively and prioritize safety measures in high-risk areas.
+By utilizing data analysis to help with policy implementation, the research aims to contribute to the reduction of school shooting casualties and foster safer educational environments.
 
-The structure of this paper is as follows: @sec-data discusses the data, including its sources, measurements, and key variables; @sec-model outlines the model used for prediction, including its justification, model validation is included in @sec-appendix; @sec-result presents the results, including summary statistics and spatial visualizations; and @sec-discussion concludes with a discussion of the findings, limitations, and recommendations for future research and policy.
+The structure of this paper is as follows: @sec-data discusses the data, including its sources, measurements, and key variables; @sec-model outlines the model used for prediction, including its justification and limitaion, model validation is included in @sec-appendix; @sec-result presents the results, including summary statistics and spatial visualizations; and @sec-discussion concludes with a discussion of the findings, limitations, and recommendations for future research and policy.
 
@@
 
-As the federal government does not consistently track school shootings, this dataset from *The Washington Post* fills a critical gap.
-It was carefully assembled using information from diverse sources, including news articles, open-source databases, law enforcement reports, and direct inquiries to schools and police departments [@washingtonpost_school_shootings].
+Since the federal government does not consistently track school shootings, this dataset from The Washington Post provides detailed information on these incidents..
 Although sources like FBI crime reports and local school incident logs were reviewed, they lack the detail and comprehensive coverage of this dataset.
@@
 
-The statistical programming language R [@citeR] is used to download, clean, analyze and model the US School Shooting Data.
-The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] .
-The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm], `arrow` [@R-arrow] and `patchwork` [@patchwork].
+The statistical programming language `R` [@citeR] is used to download, clean, analyze and model the US School Shooting Data.
+The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data].
+The following libraries in R are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm], `arrow` [@R-arrow] and `patchwork` [@patchwork].
 
@@
 @fig-casualty-by-shooting-type illustrates the distribution of casualties across different shooting types.
-The "Targeted" and "Indiscriminate" categories exhibit significantly higher casualty counts compared to other types.
+"Targeted" and "Indiscriminate" categories exhibit significantly higher casualty counts compared to other types.
 As a result, these two types are combined into one class, while the remaining types form a separate class.
@@
 
-The outcome variable for this analysis is causalities for each state in 2025. In the model, we classify the existence of casualties.
+The outcome variable for this analysis is causalities for a School Shooting Event. In the model, we classify the existence of casualties.
 
-@fig-casualtydist illustrates the distribution of casualties.
-The data is positively skewed, with the majority of cases involving fewer than 7 casualties, though there are a few school shooting incidents with exceptionally high casualty counts.
+@fig-casualtydist illustrates the distribution of casualties, highlighting a positively skewed pattern.
+The majority of incidents involve fewer than seven casualties, though a few school shootings account for exceptionally high casualty numbers.
 
@@
 @fig-Shooting_Casaulties_by_State visualizes the number of school shooting incidents in US by state for all data since 1999.
-California, Texas, Florida and North Carolina all have more than 20 school shooting incidents in the past 25 years while other states has less than 20 school shootings.
+California, Texas, Florida and North Carolina all have more than twenty school shooting incidents in the past twenty-five years while other states have less than twenty school shootings.
 
@@
 
-# Simulated dataset (replace this with your actual data)
-# Assuming `data` has a column `shooting_type`
 shooting_data <- data %>% 
@@
   mutate(
-    proportion = n / sum(n)  # Calculate proportions
+    proportion = n / sum(n),  # Calculate proportions
+    percentage = paste0(round(proportion * 100, 1), "%")  # Format as percentages
+  ) %>% 
+  arrange(desc(n))  # Arrange by frequency
+
+# Add a column to label only the top 4 categories
+shooting_data <- shooting_data %>%
+  mutate(
+    label = if_else(row_number() <= 4, percentage, "")  # Only top 4 get labels
   )
 
-# Plot Pie Chart without Text
+# Plot Pie Chart with Percentages for Top 4
 ggplot(shooting_data, aes(x = "", y = proportion, fill = shooting_type)) +
@@
   ) +
+  geom_text(
+    aes(label = label), 
+    position = position_stack(vjust = 0.5), 
+    size = 4, 
+    color = "black"
+  ) +
   theme(
@@
 
-@fig-latitudedist shows the distribution of latitude values of schools in school shooting events, ranging from approximately 20 to 50 degrees.
-Most events occur between 30 and 40 degrees, with a peak near 35 degrees.
-There are few occurrences at higher latitudes (above 45 degrees).
+@fig-latitudedist shows the distribution of latitude values of schools in school shooting events, ranging from approximately twenty to fifty degrees.
+Most events occur between thirty and forty degrees, with a peak near thirty-five degrees.
+There are few occurrences at higher latitudes (above forty-five degrees).
 
@@
 
-The Bayesian logistic regression model is implemented in **R** using the `rstanarm` package. The final model is saved as an RDS file for reproducibility and further analysis.
+The Bayesian logistic regression model is implemented in `R` using the `rstanarm` package. The final model is saved as an RDS file for reproducibility and further analysis.
 
@@
 
-The shooting types "Targeted" and "Indiscriminate" were combined into a binary variable alongside other shooting types, as exploratory data analysis in @fig-casualty-by-shooting-type revealed that these two categories are strongly associated with a higher number of casualties.
+The shooting types "targeted" and "indiscriminate" were combined into a binary variable alongside other shooting types, as exploratory data analysis in @fig-casualty-by-shooting-type illustrated that these two categories are strongly associated with a higher number of casualties.
 
@@
 
+## Overview
+
 The model results are summarized in @tbl-modelresults.
@@
 ## Intercept  
-The intercept estimate is 2.37 (95% credible interval: 0.824 to 4.02), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels.
+The intercept estimate is 2.37 (95% credible interval: 0.82 to 4.02), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels.
 This corresponds to a high baseline probability, with the credible interval excluding zero, indicating strong evidence for a nonzero probability of casualties under these conditions.
@@
   mutate(
-    exist_targeted = ifelse(prob_targeted > 0.70657, 1, 0),
-    exist_other = ifelse(prob_other > 0.70657, 1, 0)
+    exist_targeted = ifelse(prob_targeted > 0.65, 1, 0),
+    exist_other = ifelse(prob_other > 0.65, 1, 0)
   )
@@
 
-## Contributions of This Study  
 This paper utilizes a Bayesian Logistic Regression model to predict the likelihood of casualties in school shootings, focusing on the impact of shooting type and geographic latitude.
-The findings demonstrate that "Targeted" and "Indiscriminate" shootings significantly increase casualty probabilities, necessitating their classification into a unified predictive category.
+The findings demonstrate that "targeted" and "indiscriminate" shootings significantly increase casualty probabilities, necessitating their classification into a unified predictive category.
 This categorization improves the model's predictive accuracy and highlights the distinct risk these shooting types pose.
@@
 ### Geographic Targeting of Resources  
-The geographic analysis in this study reveals significant disparities in casualty risks, with southern states showing higher probabilities of school shootings resulting in casualties.
+The geographic analysis in this study shows significant disparities in casualty risks, with southern states showing higher probabilities of school shootings resulting in casualties.
 This finding emphasizes the need for tailored interventions in these regions.
@@
 ### Focus on High-Risk Shooting Types  
-The model's findings confirm that "Targeted" and "Indiscriminate" shootings are strongly associated with higher casualty rates.
+The model's findings confirm that "targeted" and "indiscriminate" shootings are strongly associated with higher casualty rates.
 Policymakers can design prevention strategies specifically aimed at these shooting types.
-For "Targeted" shootings, schools can adopt proactive threat assessment programs to identify and provide support to at-risk individuals.
-For "Indiscriminate" shootings, community-level initiatives addressing factors such as mental health, social isolation, and access to firearms can reduce the likelihood of such incidents.
+For "targeted" shootings, schools can adopt proactive threat assessment programs to identify and provide support to at-risk individuals.
+For "indiscriminate" shootings, community-level initiatives addressing factors such as mental health, social isolation, and access to firearms can reduce the likelihood of such incidents.
 
@@
 
-The Washington Post’s investigation into school shootings exemplifies a meticulous approach to data collection and analysis, aligning with best practices in data journalism and social science research.
+The Washington Post’s investigation into school shootings demonstrates a rigorous approach to data collection and analysis, adhering to established standards in data journalism and social science research.
 

### 2024-12-04 paper/paper.qmd
update graphs

@@
   labs(
-    title = "Distribution of Casualty By Shooting Type",
     x = "Shooting Type",
@@
   geom_histogram(binwidth = 1, fill = "skyblue", color = "white") + 
-  labs(title = "Distribution of Casualties", x = "Number of Casualties", y = "Frequency") +
+  labs(x = "Number of Casualties", y = "Frequency") +
   theme_minimal() +
@@
   labs(
-    title = "US School Shooting Casualties Visualization",
     x = NULL,
@@
   labs(
-    title = "Distribution of Shooting Types",
     fill = "Shooting Type"
@@
   labs(
-    title = "Distribution of Latitude of Schools in Shooting Incidents",
     x = "Latitude",

### 2024-12-04 scripts/00-simulate_data.R
update script

@@ -11,6 +11,7 @@
 #### Workspace setup ####
 library(dplyr)
 library(lubridate)
+library(arrow)
 
 #### Simulate data ####
 # Set seed for reproducibility
@@ -119,4 +120,4 @@ head(simulated_data)
 
 
 #### Save data ####
-write_csv(simulated_data, "data/00-simulated_data/simulated_data.csv")
+write_parquet(simulated_data, "data/00-simulated_data/simulated_data.parquet")

### 2024-12-04 scripts/01-test_simulated_data.R

@@ -10,8 +10,9 @@
 
 
 #### Workspace setup ####
+library(arrow)
 
-analysis_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+analysis_data <- read_parquet("data/00-simulated_data/simulated_data.parquet")
 
 # Test if the data was successfully loaded
 if (exists("analysis_data")) {

### 2024-12-04 scripts/06-model_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Models the relationship between the existence of casualty and shooting type and longtitude
+# Purpose: Models the relationship between the existence of casualty and shooting type and latitude
 # Author: Yun Chu
 # Date: 30 November 2024
 # Contact: yun.chu@mail.utoronto.ca

### 2024-12-04 other/datasheet/datasheet.qmd
rename datasheet

---
title: "Datasheet for School Shootings Dataset"
subtitle: "Data on School Shootings in the U.S. Since Columbine"
author: 
  - Yun Chu
  - Based on Washington Post Reporting
thanks: "Code and data are available at: https://github.com/washingtonpost/data-school-shootings"
date: today
abstract: "This datasheet provides information on a dataset documenting acts of gunfire at primary and secondary schools in the U.S. since Columbine. It outlines the dataset's motivation, composition, collection process, and potential uses. It aims to support informed analysis of school shootings and their impacts."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

# Datasheet for School Shootings Dataset

## Motivation

1. **Purpose:**
   - The dataset was created to identify every act of gunfire at a primary or secondary school during school hours since the Columbine High massacre on April 20, 1999. This dataset aims to quantify the impact of school shootings on children.

2. **Creators:**
   - The Washington Post team, including researchers and reporters such as John Woodrow Cox, Steven Rich, Allyson Chiu, and others.

3. **Funding:**
   - The Washington Post.

4. **Comments:**
   - The dataset was curated using Nexis, news articles, open-source databases, law enforcement reports, school websites, and direct communication with schools and police departments.

## Composition

1. **Instances Represented:**
   - Each instance represents an act of gunfire at a U.S. primary or secondary school during school hours.

2. **Total Instances:**
   - The dataset contains over 416 incidents of gun violence that fit The Post’s criteria.

3. **Completeness:**
   - This dataset is a comprehensive attempt but may not include every qualifying incident due to the lack of federal tracking.

4. **Data Content:**
   - Raw data includes variables such as date, location, injuries, casualties, and school demographic information.

5. **Labels or Targets:**
   - No specific target variable, but the dataset includes classifications of incidents.

6. **Missing Information:**
   - Some details may be unavailable or redacted.

7. **Relationships Between Instances:**
   - No explicit relationships beyond shared metadata like location or date.

8. **Data Splits:**
   - Not applicable.

9. **Errors or Noise:**
   - Efforts were made to validate data through cross-referencing.

10. **External Resources:**
   - Enrollment figures and demographic data from the U.S. Department of Education.

11. **Confidentiality:**
   - No individual-level confidential information is present.

12. **Sensitive Content:**
   - The dataset contains data on gun violence, which may be distressing.

13. **Subpopulations:**
   - Identified by school demographics.

14. **Identification of Individuals:**
   - Individuals cannot be directly identified.

15. **Sensitive Data:**
   - Includes data on school violence and student demographics.

## Collection Process

1. **Acquisition Method:**
   - Data was gathered from public records, news articles, and direct verification with schools and law enforcement.

2. **Mechanisms:**
   - Manual curation and data validation.

3. **Sampling Strategy:**
   - Incidents were selected based on predefined criteria.

4. **Contributors:**
   - Washington Post reporters and researchers.

5. **Timeframe:**
   - April 20, 1999, to Sep 6th, 2024.

6. **Ethical Review:**
   - No formal review disclosed.

7. **Source of Data:**
   - Third-party sources such as schools, law enforcement, and public databases.

8. **Notification to Individuals:**
   - Not applicable.

9. **Consent:**
   - Not applicable.

10. **Revocation Mechanism:**
    - Not applicable.

11. **Impact Analysis:**
    - Not disclosed.

## Preprocessing/Cleaning/Labeling

1. **Preprocessing:**
   - Incidents not meeting criteria (e.g., after-hours events, accidental discharges) were excluded.

2. **Raw Data Retention:**
   - Raw data is accessible via GitHub.

3. **Software Used:**
   - Details not provided.

## Uses

1. **Prior Uses:**
   - Analyzed by The Washington Post for reporting and visualization.

2. **Repository:**
   - Available on [GitHub](https://github.com/washingtonpost/data-school-shootings).

3. **Potential Tasks:**
   - Statistical analysis, policy research, educational interventions.

4. **Impact Considerations:**
   - Users should handle data with sensitivity and avoid unintended biases.

5. **Prohibited Uses:**
   - Using data to sensationalize or misrepresent school violence.

## Distribution

1. **Third-Party Sharing:**
   - Shared publicly via GitHub.

2. **Distribution Methods:**
   - GitHub repository.

3. **Availability Timeline:**
   - Dataset is continuously available.

4. **Licenses:**
   - This data is published under an Attribution-NonCommercial-ShareAlike 4.0 International (CC BY-NC-SA 4.0) license. More details are available [here](https://creativecommons.org/licenses/by-nc-sa/4.0/).

## Maintenance

1. **Support:**
   - Maintained by The Washington Post.

2. **Contact:**
   - schoolshootings@washpost.com.

3. **Erratum:**
   - Not specified.

4. **Updates:**
   - Updates are dependent on new incidents and criteria.

5. **Retention Limits:**
   - Not applicable.

6. **Versioning:**
   - Versioning policies not specified.

7. **Contributions:**
   - Not applicable.

\newpage

# References

### 2024-12-04 paper/paper.qmd
remove bold

@@
 The Post's transparency in detailing their methodology allows for reproducibility and critical assessment, key components in data journalism.
-As highlighted in *Telling Stories with Data*, reproducible workflows involve sharing data, code, and analytical steps, enabling others to replicate and validate findings [@alexander2023telling].
+As highlighted in Telling Stories with Data, reproducible workflows involve sharing data, code, and analytical steps, enabling others to replicate and validate findings [@alexander2023telling].
 
